\documentclass[master.tex]{subfiles}

\begin{document}

% ======================== TITLE ================================
\ifSubfilesClassLoaded{%
  % Standalone title block
  \begin{center}
    {\LARGE\bfseries\color{isublue}
     Chapter 3: The Do-Calculus and Identification Criteria}\\[6pt]
    {\large Theory and Methods in Causal Inference}\\[4pt]
    {\normalsize Department of Statistics, Iowa State University}
  \end{center}
  \bigskip
  \hrule height 1.2pt \relax
  \smallskip
}{%
  % Master book: chapter heading
  \chapter{The Do-Calculus and Identification Criteria}
}
% ======================== LEARNING OBJECTIVES ==================
\begin{tcolorbox}[defstyle, title={Learning Objectives}]
By the end of this chapter, students should be able to:
\begin{enumerate}
  \item Define the two intervention graph operations $\Gcal_{\overline{X}}$ and
        $\Gcal_{\underline{X}}$ and explain their causal interpretation.
  \item Apply the back-door criterion to determine whether a set $\mathbf{S}$
        identifies a causal effect, and write the back-door adjustment formula.
  \item Apply the front-door criterion in graphs where the confounder is
        unobserved but a suitable mediator exists.
  \item State all three rules of do-calculus, identify the graph condition
        that licenses each rule, and use them to prove the back-door and
        front-door formulas.
  \item State the completeness theorem of \citet{shpitser2006identification} and explain
        its practical implication: if the do-calculus cannot identify a quantity
        from the observational distribution relative to the assumed graph, then
        no purely observational method can do so without additional assumptions
        or new data.
\end{enumerate}
\end{tcolorbox}

% ---------------------------------------------------------------
% Styles and macros specific to this chapter (not yet in master).
% Placed here (after the LO box, before §1) so the definition block
% does not inject spacing between \chapter{} and the box.
% ---------------------------------------------------------------
% snode: warning-red observed node (used in §2 diagram)
\tikzset{
  snode/.style = {circle, draw=warnred, fill=warnbg, thick,
                  minimum size=7mm, font=\small\bfseries}
}
% Ancestor / Descendant set shorthands (\An, \De, \Nd defined in master.tex)
% Expectation shorthand (\E defined in master.tex)
% Remark style: thin left rule, lighter than defstyle
% (remarkstyle and remark environment are defined in causal_inference_master.tex)
% ==============================================================
\section{From d-Separation to Intervention: Intervention Graphs}
\label{w3:sec:mutilated}
% ==============================================================

In Chapter~2 we learned to read conditional independence from a DAG using
d-separation.  This chapter takes the next step: translating the graphical
language into \emph{identification formulas} --- expressions that write the
interventional distribution $P(y \mid \doop(T{=}t))$ entirely in terms of
quantities observable from data.

\begin{remark}
\textbf{How to read this chapter.}
This chapter has two pedagogical layers.
Sections~\ref{w3:sec:mutilated}--\ref{w3:sec:frontdoor} develop the two
most important concrete identification criteria --- back-door and front-door
adjustment --- using intervention graphs and direct graphical reasoning.
These sections are the core material for a first reading.

Sections~\ref{w3:sec:rules}--\ref{w3:sec:completeness} then place these
criteria inside the more general theory of the do-calculus and
identifiability.
These later sections are conceptually important, but they are more abstract
and may be read as a second pass after the concrete criteria are understood.
\end{remark}

\subsection{What Does Intervention Mean?}
\label{w3:sec:intervention-meaning}

Before introducing any graph machinery, it is worth being precise about what
the $\doop(\cdot)$ operator means and why it is not the same as conditioning.

\paragraph{Conditioning versus intervening.}
The conditional distribution $P(Y \mid T{=}t)$ describes the subpopulation of
units for whom $T$ was \emph{observed} to equal $t$.  Because treatment
assignment may be influenced by confounders, this subpopulation is not
representative of the full population.  The interventional distribution
$P(Y \mid \doop(T{=}t))$, by contrast, describes the population that would
result if $T$ were \emph{externally set} to $t$ for everyone --- removing
the natural process that determines $T$ and replacing it with the forced
value.  The two distributions coincide only when $T$ has no unmeasured
common causes with $Y$.

A simple illustration: suppose temperature $Z$ causes both ice-cream sales
$T$ and crime $Y$.
\begin{itemize}
  \item $P(Y \mid T{=}t)$ is the crime rate on days when ice-cream sales
        happen to equal $t$ --- high sales happen on hot days, which also
        drive higher crime, so the correlation is confounded by $Z$.
  \item $P(Y \mid \doop(T{=}t))$ is the crime rate we would observe if we
        \emph{fixed} sales at level $t$ by decree --- say, by regulating
        prices.  Forcing sales does not change the temperature, so crime
        remains driven only by $Z$, not by the sales level we mandated.
        If $T$ has no direct causal path to $Y$, then
        $P(Y \mid \doop(T{=}t)) = P(Y)$ for all $t$.
\end{itemize}

\paragraph{The structural basis of the do-operator.}
In the SEM framework of Chapter~1, every variable is generated by a
structural equation.  For the treatment node:
\[
  T \;=\; f_T\!\bigl(\Pa(T),\, U_T\bigr),
\]
where $\Pa(T)$ are the causal parents of $T$ and $U_T$ is exogenous noise.
An \emph{intervention} $\doop(T{=}t)$ \textbf{replaces this entire equation}
with the constant $T = t$.  The replacement has two consequences:
\begin{enumerate}
  \item $T$ is no longer influenced by its parents --- its dependence on
        $\Pa(T)$ and $U_T$ is severed.
  \item All other structural equations remain unchanged --- the rest of the
        causal mechanism is unaffected.
\end{enumerate}
This is the sense in which an intervention is a \emph{surgical} operation on
the model: it cuts out one equation and replaces it with a constant, leaving
the remainder of the system intact.

\paragraph{Graph surgery as the graphical realization.}
Because $\Pa(T)$ no longer affects $T$ after the intervention, every arrow
pointing \emph{into} $T$ in the DAG becomes meaningless and should be
removed.  The resulting graph --- the \emph{intervention graph}
$\Gcal_{\overline{T}}$ --- represents the post-intervention world.
D-separation in $\Gcal_{\overline{T}}$ therefore encodes conditional
independence in $P(\,\cdot \mid \doop(T{=}t))$, not in the original
observational distribution $P$.  This is the key link that allows
graphical reasoning to answer causal questions.

This chapter follows a \emph{concrete-to-general} order.
Sections~\ref{w3:sec:backdoor} and~\ref{w3:sec:frontdoor}
present the two most important identification strategies --- the back-door and
front-door criteria --- as self-contained results, building intuition through
worked examples.  Section~\ref{w3:sec:rules} then reveals the three rules of
do-calculus as the general algebraic engine that underpins both, and supplies
their proofs.

\subsection{The Two Graph Operations}
\label{w3:sec:operations}
Both operations modify the DAG by selectively removing edges.

\begin{defn}{Intervention Graphs $\Gcal_{\overline{X}}$ and $\Gcal_{\underline{X}}$ (Pearl's ``Mutilated Graphs'')}
Let $\Gcal = (\mathcal{V}, \mathcal{E})$ be a DAG and $X \subseteq
\mathcal{V}$ a set of nodes.
\begin{itemize}
  \item $\Gcal_{\overline{X}}$ (\emph{intervention graph}, or
        \emph{mutilated graph} in Pearl's terminology) is obtained by
        \emph{deleting all edges into} $X$: remove every arrow
        $V \to X_i$ for all $X_i \in X$.
        This represents the intervention $\doop(X{=}x)$, which
        \emph{replaces the structural equation for $X$ by the constant $x$},
        thereby severing $X$'s dependence on all its former parents.
  \item $\Gcal_{\underline{X}}$ (\emph{auxiliary observation graph}) is
        obtained by \emph{deleting all edges out of} $X$: remove every
        arrow $X_i \to V$ for all $X_i \in X$.
        This is a \emph{technical device} used in Rule~1 of the do-calculus
        to determine when an observation on $X$ can be inserted into or
        removed from a probability expression.  It does \emph{not} represent
        a probabilistic model of conditioning on $X$; it is simply the graph
        in which the relevant d-separation condition for Rule~1 must be
        checked.
\end{itemize}
\end{defn}

\noindent\textbf{Mnemonic.}
\[
  \Gcal_{\overline{X}}: \text{cut \emph{incoming} arrows into } X,
  \qquad
  \Gcal_{\underline{X}}: \text{cut \emph{outgoing} arrows from } X.
\]

Figure~\ref{w3:fig:mutilations} illustrates both intervention graph operations on the IV DAG
$\{Z \to T,\; T \to Y,\; U \to T,\; U \to Y\}$.

\begin{figure}[h]
\centering
\begin{tikzpicture}[node distance=1.8cm]
  %% ---- Original DAG ----
  \node[node]  (Z0) at (0,0)       {$Z$};
  \node[node]  (T0) at (1.9,0)     {$T$};
  \node[node]  (Y0) at (3.8,0)     {$Y$};
  \node[unode] (U0) at (2.85,-1.3) {$U$};
  \draw[edge]  (Z0) -- (T0);
  \draw[edge]  (T0) -- (Y0);
  \draw[dedge] (U0) -- (T0);
  \draw[dedge] (U0) -- (Y0);
  \node[font=\small\itshape\color{darkgrey}] at (1.9,-2.2) {Original $\Gcal$};

  %% ---- G_T-bar: delete edges INTO T ----
  \node[node]  (Z1) at (5.8,0)      {$Z$};
  \node[node]  (T1) at (7.7,0)      {$T$};
  \node[node]  (Y1) at (9.6,0)      {$Y$};
  \node[unode] (U1) at (8.65,-1.3)  {$U$};
  \draw[xedge] (Z1) -- (T1);
  \draw[edge]  (T1) -- (Y1);
  \draw[xedge] (U1) -- (T1);
  \draw[dedge] (U1) -- (Y1);
  \node[font=\small\itshape\color{darkgrey}] at (7.7,-2.2)
    {$\Gcal_{\overline{T}}$: delete arrows \emph{into} $T$};

  %% ---- G_T-underline: delete edges OUT of T ----
  \node[node]  (Z2) at (11.6,0)     {$Z$};
  \node[node]  (T2) at (13.5,0)     {$T$};
  \node[node]  (Y2) at (15.4,0)     {$Y$};
  \node[unode] (U2) at (14.45,-1.3) {$U$};
  \draw[edge]  (Z2) -- (T2);
  \draw[xedge] (T2) -- (Y2);
  \draw[dedge] (U2) -- (T2);
  \draw[dedge] (U2) -- (Y2);
  \node[font=\small\itshape\color{darkgrey}] at (13.5,-2.2)
    {$\Gcal_{\underline{T}}$: delete arrows \emph{out of} $T$};
\end{tikzpicture}
\caption{The two graph operations on the IV DAG.  Faded arrows have been
removed.  $\Gcal_{\overline{T}}$ models external intervention $\doop(T{=}t)$:
the back-door path $T\leftarrow U \to Y$ is severed.
$\Gcal_{\underline{T}}$ deactivates $T$'s influence on its descendants,
used by Rule~1 to add or delete $T$ from a conditioning set.}
\label{w3:fig:mutilations}
\end{figure}

\subsection{Why the Intervention Graph Works}
\label{w3:sec:mutilation-principle}
The causal interpretation rests on a key structural insight:

\begin{tcolorbox}[defstyle, title={The Intervention Principle}]
In the original DAG $\Gcal$, the node $T$ is determined by its parents
$\Pa(T)$ via the structural equation $T = f_T(\Pa(T), \varepsilon_T)$.
Intervening on $T$ removes the structural mechanism that normally determines
$T$; therefore the parents of $T$ no longer influence it.
The intervention $\doop(T{=}t)$ \emph{replaces this equation} with the
constant $T = t$, making $T$ independent of all its former parents.
Graphically, this severs every arrow into $T$.  The remaining structure of
the graph is unchanged: all other structural equations still hold.

A d-separation statement in $\Gcal_{\overline{T}}$ is therefore a
conditional independence statement in the \emph{post-intervention}
distribution $P(\,\cdot \mid \doop(T{=}t))$, not in the observational
distribution $P$.  Section~\ref{w3:sec:rules} develops the three do-calculus
rules that make this correspondence precise and supplies the proofs of the
formulas introduced in Sections~\ref{w3:sec:backdoor}
and~\ref{w3:sec:frontdoor}.
\end{tcolorbox}

% ==============================================================
\section{The Back-Door Criterion}
\label{w3:sec:backdoor}
% ==============================================================

The back-door criterion is the most widely used identification strategy.
It gives a simple graphical condition under which conditioning on an
observed set $\mathbf{S}$ is sufficient to identify the causal effect of
$T$ on $Y$.  The formula it produces --- standardization over $\mathbf{S}$
--- is computable directly from the observed data distribution;
Section~\ref{w3:sec:proofs} supplies the do-calculus proof.

\subsection{Motivation: When the Confounder Is Unobserved}
\label{w3:sec:backdoor-motiv}

If all common causes of $T$ and $Y$ were observed, identification would be trivial: adjust
for all of them.  The interesting case --- and the one the back-door criterion is designed
for --- is when some confounders are \emph{unobserved}, yet an observed set $\mathbf{S}$
can still block every spurious path.

Consider the following DAG.  A drug $T$ affects recovery $Y$.
Socioeconomic status $U$ is an unobserved confounder ($U \to T$, $U \to Y$),
but $U$ also affects an observed proxy $W$ ($U \to W$), which in turn affects $Y$
($W \to Y$).

\begin{center}
\begin{tikzpicture}[node distance=1.8cm]
  \node[unode]  (U) at (1.8,  0)   {$U$};
  \node[node]   (W) at (3.6,  0)   {$W$};
  \node[node]   (T) at (0,   -1.4) {$T$};
  \node[node]   (Y) at (3.6, -1.4) {$Y$};
  \draw[edge]   (T) -- (Y);
  \draw[dedge]  (U) -- (T);
  \draw[dedge]  (U) -- (W);
  \draw[edge]   (W) -- (Y);
\end{tikzpicture}
\end{center}

\noindent
The only back-door path is $T \leftarrow U \to W \to Y$.  Although $U$ is unobserved
and cannot be adjusted for directly, $W$ sits on this path and \emph{is} observed.
Conditioning on $W$ blocks the path at the chain node $W$, and $W$ is not a descendant
of $T$, so $\mathbf{S} = \{W\}$ satisfies the back-door criterion.  The causal effect is
therefore identified even though $U$ is never measured.

This is the key insight: the criterion does not require adjusting for the confounders
themselves, only for an observed set that \emph{intercepts} every back-door path.

\subsection{Definition and Theorem}
\label{w3:sec:backdoor-def}
\begin{defn}{Back-Door Criterion}
\label{w3:def:backdoor}
A set of observed variables $\mathbf{S}$ satisfies the \emph{back-door
criterion} for the causal effect of $T$ on $Y$ in $\Gcal$ if:
\begin{enumerate}
  \item No node in $\mathbf{S}$ is a descendant of $T$.
  \item $\mathbf{S}$ blocks every \emph{back-door path} from $T$ to $Y$,
        i.e., every path that begins with an arrow pointing \emph{into}
        $T$ (a path of the form $T \leftarrow \cdots Y$).
\end{enumerate}
\end{defn}

\begin{remark}
The terminology reflects the geometry of the graph relative to $T$.
A \emph{back-door path} enters $T$ from behind --- it begins with an arrow
pointing \emph{into} $T$ --- and represents a confounding route through
which $T$ and $Y$ become spuriously associated without any causal mechanism
from $T$ to $Y$.  By contrast, the directed causal paths
$T \to \cdots \to Y$ leave $T$ through the \emph{front} and carry the
genuine effect.  The front-door criterion of
Section~\ref{w3:sec:frontdoor} is named for exactly this complementary
idea: it identifies the causal effect by intercepting the \emph{front-door}
(causal) paths at a mediator $M$, rather than blocking the back-door
(confounding) paths directly.
Standing at $T$, confounders sneak in through the back door; the causal
effect walks out through the front.
\end{remark}

\begin{remark}
The adjustment set $\mathbf{S}$ here plays the same role as the
conditioning set in the d-separation notation of Chapter~2.
Specifically, condition~2 is equivalent to requiring
$(Y \indep T \mid \mathbf{S})_{\Gcal_{\underline{T}}}$:
$\mathbf{S}$ d-separates $T$ and $Y$ in the graph $\Gcal_{\underline{T}}$
obtained by \emph{deleting all arrows \textbf{out of} $T$}.  Removing
$T$'s outgoing edges eliminates the causal paths $T \to \cdots \to Y$,
leaving only back-door paths between $T$ and $Y$; condition~2 therefore
reduces to checking d-separation in $\Gcal_{\underline{T}}$.  (The
intervention graph $\Gcal_{\overline{T}}$, which deletes arrows
\emph{into} $T$, is the wrong graph for this check: it preserves
$T \to Y$ as an open path, so $(Y \indep T \mid \mathbf{S})_{\Gcal_{\overline{T}}}$
fails even for valid adjustment sets.)
The back-door criterion adds two constraints that $\mathbf{S}$ in
Chapter~2 did not carry: $\mathbf{S}$ must consist of \emph{observed}
variables only, and it must contain no descendant of $T$.  Every valid
adjustment set $\mathbf{S}$ is therefore a valid d-separator in
$\Gcal_{\underline{T}}$, but the converse does not hold --- a set that
d-separates $T$ and $Y$ in $\Gcal_{\underline{T}}$ may still fail the
back-door criterion if it contains a descendant of $T$ or an unobserved
variable.
\end{remark}

\begin{example}{A Single Confounder}
\label{w3:ex:singleconf}
Consider the DAG $T \leftarrow C \to Y$ with $T \to Y$, where $C$ is an
observed confounder.  We wish to identify $P(y \mid \doop(T{=}t))$.

\begin{center}
\begin{tikzpicture}[node distance=1.8cm]
  \node[node] (C) at (1.8, 0)   {$C$};
  \node[node] (T) at (0,  -1.4) {$T$};
  \node[node] (Y) at (3.6,-1.4) {$Y$};
  \draw[edge] (T) -- (Y);
  \draw[edge] (C) -- (T);
  \draw[edge] (C) -- (Y);
\end{tikzpicture}
\end{center}

\noindent
There is one back-door path: $T \leftarrow C \to Y$ (an arrow points into
$T$).  Does $\mathbf{S} = \{C\}$ satisfy the back-door criterion?
\begin{enumerate}
  \item $C$ is not a descendant of $T$. \checkmark
  \item $\{C\}$ blocks $T \leftarrow C \to Y$ (fork at $C$). \checkmark
\end{enumerate}
The criterion is satisfied, and the back-door formula gives:
\[
  P(y \mid \doop(T{=}t)) = \sum_c P(y \mid t, c)\,P(c).
\]
This is the simplest possible application: one confounder, one back-door
path, one adjustment variable.
\end{example}

\begin{example}{Two Back-Door Paths: Both Must Be Blocked}
\label{w3:ex:twopaths}
Condition~2 is a \emph{conjunction}: $\mathbf{S}$ must block
\emph{every} back-door path, not merely the most visible one.  This
example shows that blocking each path requires a separate witness in
$\mathbf{S}$, and that blocking one while leaving the other open still
yields a biased estimator.

\medskip
\noindent\textbf{Graph.}
Consider a treatment $T$, outcome $Y$, observed variables $C_1$ and
$C_2$, and an unobserved confounder $U$:

\begin{center}
\begin{tikzpicture}[node distance=1.6cm]
  \node[node]  (C1) at (0.0, 1.5) {$C_1$};
  \node[unode] (U)  at (2.2, 1.5) {$U$};
  \node[node]  (C2) at (4.4, 1.5) {$C_2$};
  \node[node]  (T)  at (0.0, 0)   {$T$};
  \node[node]  (Y)  at (4.4, 0)   {$Y$};
  \draw[edge]  (C1) -- (T);
  \draw[edge]  (C1) -- (Y);
  \draw[dedge] (U)  -- (T);
  \draw[dedge] (U)  -- (C2);
  \draw[edge]  (C2) -- (Y);
  \draw[edge]  (T)  -- (Y);
\end{tikzpicture}
\end{center}

\noindent
There are exactly two back-door paths from $T$ to $Y$:
\begin{enumerate}
  \item $T \leftarrow C_1 \to Y$\quad (direct confounding by the observed $C_1$),
  \item $T \leftarrow U \to C_2 \to Y$\quad (indirect confounding via the
        unobserved $U$, which routes through the observed proxy $C_2$).
\end{enumerate}

\medskip
\noindent\textbf{Does $\mathbf{S} = \{C_1\}$ work?}
\begin{itemize}
  \item Condition~1: $C_1$ is not a descendant of $T$. \checkmark
  \item Condition~2: $\{C_1\}$ blocks path~1 (fork blocked at $C_1$) \checkmark,
        but leaves path~2 open ($C_1$ is not on the path
        $T \leftarrow U \to C_2 \to Y$, and $C_2$ is unconditioned). $\times$
\end{itemize}
\textbf{Fails}: the residual confounding via $U \to C_2 \to Y$ is uncontrolled.

\medskip
\noindent\textbf{Does $\mathbf{S} = \{C_2\}$ work?}
\begin{itemize}
  \item Condition~1: $C_2$ is not a descendant of $T$. \checkmark
  \item Condition~2: $\{C_2\}$ blocks path~2 (chain blocked at $C_2$) \checkmark,
        but leaves path~1 open ($C_2$ is not on $T \leftarrow C_1 \to Y$,
        and $C_1$ is unconditioned). $\times$
\end{itemize}
\textbf{Fails}: the direct confounding via $C_1$ is uncontrolled.

\medskip
\noindent\textbf{Does $\mathbf{S} = \{C_1, C_2\}$ work?}
\begin{itemize}
  \item Condition~1: neither $C_1$ nor $C_2$ is a descendant of $T$. \checkmark
  \item Condition~2: $C_1$ blocks path~1; $C_2$ blocks path~2. \checkmark
\end{itemize}
\textbf{Succeeds.}  The back-door formula gives:
\[
  P(y \mid \doop(T{=}t))
  \;=\; \sum_{c_1,\,c_2} P(y \mid t,\,c_1,\,c_2)\,P(c_1,\,c_2).
\]

\medskip
\noindent\textbf{Remark.}
The unobserved $U$ never appears in the formula: what matters is
that $C_2$ sits on the back-door path between $T$ and $Y$ (specifically,
on the segment $U \to C_2 \to Y$) and is observed.  Conditioning on $C_2$
blocks the path at the chain node $C_2$, just as conditioning on $C_1$
blocks it at the fork node $C_1$.  The identification strategy exploits
the \emph{position} of the observed variables on the back-door paths, not
direct measurement of the confounders themselves.
\end{example}

\begin{warn}{Why Condition~1 Is Essential: A Worked Counterexample}
Condition~1 rules out conditioning on any descendant of $T$.  To see why,
consider a drug $T$ that affects recovery $Y$ via two routes: a direct effect
$T \to Y$ and an indirect effect through blood pressure $M$ ($T \to M \to Y$).
Suppose also that an unobserved confounder $U$ affects both $T$ and $Y$.

\begin{center}
\begin{tikzpicture}[node distance=1.8cm]
  \node[unode] (U) at (1.8,  0)   {$U$};
  \node[node]  (T) at (0,   -1.4) {$T$};
  \node[node]  (M) at (1.8, -1.4) {$M$};
  \node[node]  (Y) at (3.6, -1.4) {$Y$};
  \draw[edge]  (T) -- (M);
  \draw[edge]  (M) -- (Y);
  \draw[edge]  (T) to[bend right=30] (Y);
  \draw[dedge] (U) -- (T);
  \draw[dedge] (U) -- (Y);
\end{tikzpicture}
\end{center}

\noindent
The back-door path is $T \leftarrow U \to Y$.  Now suppose we naively try
$\mathbf{S} = \{M\}$, hoping that conditioning on $M$ will block this path.
It does not: $M$ lies on the \emph{causal} path $T \to M \to Y$, not on the
back-door path $T \leftarrow U \to Y$.  Conditioning on $M$ therefore fails
for two distinct reasons:
\begin{enumerate}
  \item It violates Condition~1: $M$ is a descendant of $T$.
  \item It \emph{blocks causal information}: the stratum-specific association
        $P(y \mid t, m)$ is computed holding $M$ fixed, so it cannot see the
        effect of $T$ that flows through $M$.  The result $\sum_m P(y \mid t,m)P(m)$
        estimates the direct effect $T \to Y$ only, not the total causal effect.
\end{enumerate}
The correct adjustment for the total effect uses an observed set that blocks the
back-door path $T \leftarrow U \to Y$ \emph{without touching} the causal paths.
If $U$ is unobserved and no other observed variable sits on the back-door path,
the total causal effect is \emph{not identified} by back-door adjustment in this graph.
\end{warn}

\begin{warn}{Why Condition~2 Can Be Violated by a Collider: M-Bias}
Condition~2 requires $\mathbf{S}$ to \emph{block} every back-door path.
A subtler failure mode than including a descendant of $T$ is including a
variable that is a \emph{collider} on a back-door path: conditioning on a
collider \emph{opens} that path rather than blocking it, introducing
confounding where none existed before.

Consider the following DAG. The treatment $T$ has a direct effect on $Y$.
Two unobserved variables $U_1$ and $U_2$ share a common \emph{observed}
descendant $C$ (e.g., a baseline characteristic recorded before treatment),
with $U_1$ also affecting $T$ and $U_2$ also affecting $Y$.

\begin{center}
\begin{tikzpicture}[node distance=1.8cm]
  \node[node]  (T) at (0,    0)    {$T$};
  \node[node]  (C) at (2.0,  1.3)  {$C$};
  \node[node]  (Y) at (4.0,  0)    {$Y$};
  \node[unode] (U1) at (0.8,  2.4)  {$U_1$};
  \node[unode] (U2) at (3.2,  2.4)  {$U_2$};
  \draw[edge]  (T) -- (Y);
  \draw[dedge] (U1) -- (T);
  \draw[dedge] (U1) -- (C);
  \draw[dedge] (U2) -- (C);
  \draw[dedge] (U2) -- (Y);
\end{tikzpicture}
\end{center}

\noindent
The only back-door path from $T$ to $Y$ is
$T \leftarrow U_1 \to C \leftarrow U_2 \to Y$.
Node $C$ is a \emph{collider} on this path.  By the d-separation rules of
Chapter~2, a collider \emph{blocks} the path by default (without any
conditioning), so the causal effect of $T$ on $Y$ is already identified
with $\mathbf{S} = \varnothing$: $P(y \mid \doop(t)) = P(y \mid t)$.

Now suppose a researcher naively includes $C$ in the adjustment set,
reasoning that $C$ is a pretreatment covariate and ``more adjustment is
safer.''  This reasoning is wrong.  Adding $C$ to $\mathbf{S}$
\emph{activates} the collider, opening the path
$T \leftarrow U_1 \to C \leftarrow U_2 \to Y$.  The back-door path, which
was dormant, now transmits a spurious association between $T$ and $Y$
through the unobserved variables $U_1$ and $U_2$.  The set $\{C\}$
therefore fails Condition~2: it does not block the back-door path; it
creates one.  The resulting estimator $\sum_c P(y \mid t, c)\,P(c)$ is
\emph{biased} for the causal effect.

This phenomenon is called \emph{M-bias} because the graph has an M-shaped
skeleton ($T \leftarrow U_1 \to C \leftarrow U_2 \to Y$).  The practical
lesson is that Condition~1 alone (``include all pretreatment variables'')
is not sufficient; every candidate adjustment variable must also be
verified not to open a previously closed back-door path.
\end{warn}

\begin{thm}{Back-Door Adjustment Formula \citep{pearl1993comment}}
\label{w3:thm:backdoor}
If $\mathbf{S}$ satisfies the back-door criterion for the effect of $T$ on
$Y$ in $\Gcal$, and
\begin{equation}
  P(T{=}t \mid \mathbf{S}{=}\mathbf{s}) \;>\; 0
  \quad \text{for all } \mathbf{s} \text{ with } P(\mathbf{S}{=}\mathbf{s}) > 0,
  \label{w3:eq:positivity}
\end{equation}
then:
\begin{equation}
  P\!\left(y \mid \doop(T{=}t)\right)
  \;=\; \int P(y \mid T{=}t,\;\mathbf{S}{=}\mathbf{s})\,dP(\mathbf{s}).
  \label{w3:eq:backdoor}
\end{equation}
Equivalently, for any measurable function $h$ with $\E|h(Y)| < \infty$,
\begin{equation}
  \E\!\left[h(Y) \mid \doop(T{=}t)\right]
  \;=\; \E_{\mathbf{S}}\!\Bigl[\E\bigl[h(Y) \mid T{=}t,\;\mathbf{S}\bigr]\Bigr].
  \label{w3:eq:backdoor-exp}
\end{equation}
The interventional distribution is identified by integrating the
conditional distribution $P(y \mid t, \mathbf{s})$ with respect to the
\emph{marginal} distribution of $\mathbf{S}$.  The iterated-expectation
form~\eqref{w3:eq:backdoor-exp} makes this structure transparent: the
outer expectation averages over the population distribution of $\mathbf{S}$,
while the inner expectation computes the stratum-specific association --- which,
within each stratum, is causal because all back-door paths are blocked.
When $\mathbf{S}$ is discrete, the integral reduces to
$\sum_{\mathbf{s}} P(y \mid t, \mathbf{s})\,P(\mathbf{s})$.
\end{thm}

\begin{remark}
Condition~\eqref{w3:eq:positivity} is the \emph{positivity} (or
\emph{overlap}) assumption.  It requires every value of the treatment
$T{=}t$ to be observable within each stratum of $\mathbf{S}$: if some
stratum $\mathbf{s}$ never receives treatment $t$ in the data, the
conditional distribution $P(y \mid t, \mathbf{s})$ is undefined, and the
right-hand side of~\eqref{w3:eq:backdoor} cannot be computed.
Positivity is a \emph{data-support} condition, not a graphical one: the
back-door criterion is a property of the DAG, but positivity is a property
of the joint distribution $P$.  Both must hold simultaneously for the
adjustment formula to be well-defined.  When positivity fails on a set of
measure zero, the formula still identifies $P(y \mid \doop(t))$ because
the offending strata contribute zero mass to the outer integral; the
concern is practical failure (sparse or empty cells) rather than a
theoretical identification failure.
\end{remark}

The derivation uses Rules~2 and~3 of the do-calculus; the complete
three-step proof is given in Section~\ref{w3:sec:proofs}.

\subsection{The Adjustment Formula: Standardization}
\label{w3:sec:standardization}
Equation~\eqref{w3:eq:backdoor} is also known as the \emph{standardization
formula} or, in the epidemiological literature, the \emph{g-formula}
\citep{robins1986new}.  The g-formula was originally derived for the single
time-point case $K=1$; its multi-period generalization to dynamic treatment
regimes --- where treatment and covariates are both time-varying --- is the
subject of Chapter~13.  The interpretation of the formula is important:

\begin{itemize}
  \item $P(y \mid t, \mathbf{s})$ is the stratum-specific conditional
        distribution of $Y$ given $T{=}t$ and $\mathbf{S}{=}\mathbf{s}$.
        Within each stratum, $T$ is not confounded by any back-door path
        (all are blocked by $\mathbf{S}$), so this conditional distribution
        \emph{is} causal.
  \item The outer integration $\int \cdots\, dP(\mathbf{s})$ weights the
        strata by the \emph{population marginal} distribution of
        $\mathbf{S}$, rather than the treatment-conditional distribution
        $dP(\mathbf{s} \mid t)$ that a naive analysis would use.  The
        iterated-expectation form~\eqref{w3:eq:backdoor-exp} makes this
        re-weighting explicit: $\E_{\mathbf{S}}[\cdots]$ is expectation
        under $P(\mathbf{S})$, not $P(\mathbf{S} \mid T{=}t)$.
  \item In a randomized experiment, $\mathbf{S} = \varnothing$ satisfies
        the back-door criterion (no confounding), and
        \eqref{w3:eq:backdoor} reduces to $P(y \mid \doop(t)) = P(y \mid t)$:
        the observational and interventional distributions coincide.
\end{itemize}

\begin{example}{Back-Door Adjustment: A Numerical Walkthrough}
\label{w3:ex:backdoor-numerical}
We now trace the adjustment formula~\eqref{w3:eq:backdoor} through explicit
numbers.  Use the single-confounder graph of Example~\ref{w3:ex:singleconf}
($T \leftarrow C \to Y$, $T \to Y$) with all variables binary:
$C$ = disease severity ($0$ = mild, $1$ = severe),
$T$ = treatment ($0$ = no, $1$ = yes),
$Y$ = recovery ($0$ = no, $1$ = yes).

\medskip
\noindent\textbf{Data-generating distribution.}
Severity is equally prevalent, $P(C{=}1) = 0.5$, and \emph{sicker patients
are preferentially treated}: $P(T{=}1 \mid C{=}0) = 0.20$ and
$P(T{=}1 \mid C{=}1) = 0.80$.  The outcome probabilities are:

\medskip
\begin{center}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{@{\hspace{6pt}}ccc@{\hspace{20pt}}cc@{\hspace{6pt}}}
\toprule
$C$ & $P(C)$ & $P(T{=}1\mid C)$ & $P(Y{=}1\mid T{=}0,\,C)$ & $P(Y{=}1\mid T{=}1,\,C)$ \\
\midrule
0 & 0.50 & 0.20 & 0.10 & 0.40 \\
1 & 0.50 & 0.80 & 0.50 & 0.80 \\
\bottomrule
\end{tabular}
\end{center}

\medskip\noindent
The treatment effect within each stratum is
$P(Y{=}1\mid T{=}1,C{=}c) - P(Y{=}1\mid T{=}0,C{=}c) = 0.30$ for both
$c=0$ and $c=1$, so the true average causal effect is $\mathrm{ACE} = 0.30$.

\medskip
\noindent\textbf{Step 1: The naïve (unadjusted) estimand.}
Because $P(T{=}1) = 0.5$, the marginal outcome rates are:
\begin{align*}
  P(Y{=}1 \mid T{=}1)
  &= \frac{P(Y{=}1,\,T{=}1)}{P(T{=}1)}
   = \frac{0.40\times 0.20\times 0.5 \;+\; 0.80\times 0.80\times 0.5}{0.5}
   = \frac{0.04 + 0.32}{0.5} = 0.72,\\[4pt]
  P(Y{=}1 \mid T{=}0)
  &= \frac{0.10\times 0.80\times 0.5 \;+\; 0.50\times 0.20\times 0.5}{0.5}
   = \frac{0.04 + 0.05}{0.5} = 0.18.
\end{align*}
Na\"{i}ve difference: $0.72 - 0.18 = \mathbf{0.54}$.

\medskip
\noindent\textbf{Step 2: The back-door adjusted estimand.}
Since $\{C\}$ satisfies the back-door criterion (no descendant of $T$, blocks
$T \leftarrow C \to Y$), formula~\eqref{w3:eq:backdoor} applies.  Weighting
the stratum-specific outcome rates by the \emph{population} marginal
$P(C)$ rather than by $P(C \mid T)$:
\begin{align*}
  P\!\left(Y{=}1 \mid \doop(T{=}1)\right)
  &= P(Y{=}1\mid T{=}1,C{=}0)\,P(C{=}0)
    + P(Y{=}1\mid T{=}1,C{=}1)\,P(C{=}1)\\
  &= 0.40\times 0.5 \;+\; 0.80\times 0.5 \;=\; 0.60,\\[4pt]
  P\!\left(Y{=}1 \mid \doop(T{=}0)\right)
  &= 0.10\times 0.5 \;+\; 0.50\times 0.5 \;=\; 0.30.
\end{align*}
Adjusted difference: $0.60 - 0.30 = \mathbf{0.30}$ --- matching the true ACE.

\medskip
\noindent\textbf{Diagnosis of the bias.}
The na\"{i}ve estimate ($0.54$) overstates the causal effect by $0.24$.
The source is the severity--treatment association: treated patients are
predominantly severe ($C{=}1$), who recover at higher rates \emph{even
without treatment} ($0.50$ vs.\ $0.10$).  Marginal comparison therefore
conflates the treatment benefit with the baseline severity difference.
The back-door formula removes this confounding by imposing the
\emph{same} covariate distribution $P(C)$ on both potential worlds,
recovering the true ACE.

\medskip
\noindent\textbf{What the formula is not.}
Note that the adjusted formula is \emph{not}
$\sum_c P(Y{=}1\mid T{=}1,C{=}c)\,P(C{=}c \mid T{=}1)$: that is just
$P(Y{=}1 \mid T{=}1) = 0.72$ again.  The entire force of the adjustment
lies in replacing the treatment-conditional distribution $P(C \mid T)$
with the population marginal $P(C)$, as equation~\eqref{w3:eq:backdoor}
requires.
\end{example}

\subsection{Application to the Education--Earnings DAG}
\label{w3:sec:education-ex}
\begin{example}{Back-Door Identification in the Education Example}
\label{w3:ex:education}
This is the Education--Earnings DAG analyzed in full in
Section~\ref{w2:sec:education} of Chapter~2, where Steps~1--3 of the
identification pipeline were carried out.  We now complete Step~4: applying the back-door criterion to
derive the identification formula.  The DAG has edges $N \to E$, $B \to E$,
$B \to Y$, $E \to Y$, where $N$ = neighborhood, $B$ = family background,
$E$ = education, $Y$ = earnings.  We wish to identify
$P(y \mid \doop(E{=}e))$.

\begin{center}
\begin{tikzpicture}[node distance=1.8cm]
  \node[node] (N) at (0, 0)     {$N$};
  \node[node] (B) at (2.0, 0)   {$B$};
  \node[node] (E) at (1.0,-1.4) {$E$};
  \node[node] (Y) at (3.0,-1.4) {$Y$};
  \draw[edge] (E) -- (Y);
  \draw[edge] (N) -- (E);
  \draw[-{Stealth[length=5pt]}, thick, color=warnred, dashed] (B) -- (E);
  \draw[-{Stealth[length=5pt]}, thick, color=warnred, dashed] (B) -- (Y);
  \node[font=\scriptsize\itshape\color{darkgrey}, above=0.05cm of N] {neighborhood};
  \node[font=\scriptsize\itshape\color{darkgrey}, above=0.05cm of B] {background};
  \node[font=\scriptsize\itshape\color{darkgrey}, below=0.05cm of E] {education};
  \node[font=\scriptsize\itshape\color{darkgrey}, below=0.05cm of Y] {earnings};
  \draw[-{Stealth[length=5pt]}, thick, color=warnred, dashed]
    (0.0,-2.3) -- (0.6,-2.3)
    node[anchor=west, font=\small\color{warnred}]
    {\; back-door path $E \leftarrow B \to Y$};
\end{tikzpicture}
\end{center}

\medskip\noindent
\textbf{Back-door paths from $E$ to $Y$:}
The only back-door path is $E \leftarrow B \to Y$ (an arrow points into
$E$ from $B$, then continues to $Y$).

\medskip\noindent
\textbf{Does $\{B\}$ satisfy the criterion?}
\begin{enumerate}
  \item $B$ is not a descendant of $E$. \checkmark
  \item $\{B\}$ blocks $E \leftarrow B \to Y$ (fork at $B$,
        conditioning closes it). \checkmark
\end{enumerate}
By Theorem~\ref{w3:thm:backdoor}:
\[
  P(y \mid \doop(E{=}e)) = \sum_b P(y \mid e, b)\,P(b).
\]

\medskip\noindent
\textbf{Does $\{N\}$ satisfy the criterion?}
$N$ is not a descendant of $E$, but $\{N\}$ does \emph{not} block
$E \leftarrow B \to Y$ (the fork at $B$ remains open).  So $\{N\}$ fails
condition~2 and cannot be used for adjustment.

\medskip\noindent
\textbf{Does $\{N, B\}$ satisfy the criterion?}
Both conditions hold: neither $N$ nor $B$ is a descendant of $E$, and
$\{N,B\}$ blocks the only back-door path.  So $\{N,B\}$ also yields a
valid (though redundant) adjustment formula.
\end{example}

\begin{remark}
The Education example illustrates three important points about the set of
valid adjustment sets.

\textbf{(a) Multiple valid sets may exist, and they identify the same
quantity.}  Both $\{B\}$ and $\{N,B\}$ satisfy the back-door criterion
here.  Any such set yields the same interventional distribution
$P(y \mid \doop(E{=}e))$; the identification target is unique even though
the adjustment sets are not.

\textbf{(b) Supersets of a valid adjustment set are not automatically
valid.}  Adding a variable to a valid set can violate Condition~2 if that
variable is a collider on some back-door path (as in the M-bias example
above).  In this particular graph $\{N,B\}$ happens to remain valid
because $N$ is not a collider on any back-door path and creates no new
open path.  But this must be verified from the graph, not assumed.

\textbf{(c) Smaller valid sets are often preferable in practice.}  Every
variable added to $\mathbf{S}$ must also satisfy positivity
(condition~\eqref{w3:eq:positivity}).  Adjusting for $N$ in addition to
$B$ requires $P(E{=}e \mid B{=}b, N{=}n) > 0$ for all cells $(b,n)$, a
stricter requirement than $P(E{=}e \mid B{=}b) > 0$ alone.  Sparse or
empty cells in high-dimensional strata inflate variance and cause practical
positivity failure.  Adjusting for fewer variables --- the minimal
sufficient set --- is generally more efficient and more robust.
\end{remark}

% ==============================================================
\section{The Front-Door Criterion}
\label{w3:sec:frontdoor}
% ==============================================================

The back-door criterion requires that all confounders be observed.  In
many applications --- including the IV setting --- the confounder $U$ is
unobserved and no back-door adjustment is possible.  The front-door
criterion provides identification in a different configuration: when there
exists a measured \emph{mediator} $M$ on the causal path from $T$ to $Y$
with specific independence properties.

\subsection{The Configuration and Criterion}
\label{w3:sec:frontdoor-def}
\begin{defn}{Front-Door Criterion}
\label{w3:def:frontdoor}
A set of variables $\mathbf{M}$ satisfies the \emph{front-door criterion}
for the effect of $T$ on $Y$ in $\Gcal$ if:
\begin{enumerate}
  \item $\mathbf{M}$ intercepts all directed paths from $T$ to $Y$
        (all causal paths pass through $\mathbf{M}$).
  \item There are no unblocked back-door paths from $T$ to $\mathbf{M}$
        (the effect of $T$ on $\mathbf{M}$ is not confounded).
  \item All back-door paths from $\mathbf{M}$ to $Y$ are blocked by $T$
        (conditioning on $T$ closes all confounding of the
        $\mathbf{M} \to Y$ effect).
\end{enumerate}
\end{defn}

\begin{example}{A Single Mediator with Unobserved Confounding}
\label{w3:ex:frontdoorconds}
Consider the DAG $T \to M \to Y$ with $U \to T$ and $U \to Y$, where
$M$ is observed and $U$ is unobserved.

\begin{center}
\begin{tikzpicture}[node distance=1.8cm]
  \node[node]  (T) at (0,0)      {$T$};
  \node[node]  (M) at (2.0,0)    {$M$};
  \node[node]  (Y) at (4.0,0)    {$Y$};
  \node[unode] (U) at (2.0,-1.4) {$U$};
  \draw[edge]  (T) -- (M);
  \draw[edge]  (M) -- (Y);
  \draw[dedge] (U) -- (T);
  \draw[dedge] (U) -- (Y);
\end{tikzpicture}
\end{center}

\noindent Does $M$ satisfy the front-door criterion?
\begin{enumerate}
  \item \emph{$M$ intercepts all directed paths from $T$ to $Y$.}
        The only directed path is $T \to M \to Y$, which passes through
        $M$. \checkmark
  \item \emph{No unblocked back-door paths from $T$ to $M$.}
        The only path from $T$ to $M$ that is not the direct arrow
        $T \to M$ would have to go through $U$, but there is no edge
        $U \to M$, so no back-door path exists. \checkmark
  \item \emph{All back-door paths from $M$ to $Y$ are blocked by $T$.}
        The only back-door path from $M$ to $Y$ is
        $M \leftarrow T \leftarrow U \to Y$.
        Conditioning on $T$ blocks this path (chain at $T$). \checkmark
\end{enumerate}
All three conditions hold, so the front-door formula applies:
\[
  P(y \mid \doop(T{=}t))
  = \sum_m P(m \mid t)\sum_{t'} P(y \mid t', m)\,P(t').
\]
\end{example}

\begin{example}{When the Front-Door Criterion Fails}
\label{w3:ex:frontdoorfail}
Each condition of Definition~\ref{w3:def:frontdoor} is load-bearing.
We illustrate what breaks when exactly one condition fails, by modifying
the prototypical front-door graph in two separate ways.

\medskip
\noindent\textbf{Case (a): Condition~2 fails --- the $T \to M$ link is confounded.}

Start with the prototypical graph and add the edge $U \to M$, so the
unobserved confounder also directly affects the mediator:

\begin{center}
\begin{tikzpicture}[node distance=1.6cm]
  \node[node]  (T) at (0, 0)     {$T$};
  \node[node]  (M) at (2.2, 0)   {$M$};
  \node[node]  (Y) at (4.4, 0)   {$Y$};
  \node[unode] (U) at (1.1,-1.3) {$U$};
  \draw[edge]  (T) -- (M);
  \draw[edge]  (M) -- (Y);
  \draw[dedge] (U) -- (T);
  \draw[dedge, color=warnred] (U) -- (M)
    node[midway, below right, font=\scriptsize\color{warnred}] {new};
  \draw[dedge] (U) -- (Y);
\end{tikzpicture}
\end{center}

\noindent
Condition~1 still holds ($M$ intercepts all directed paths $T \to M \to Y$).
But condition~2 now \textbf{fails}: the path $T \leftarrow U \to M$ is an
unblocked back-door path from $T$ to $M$.  Because $U$ is unobserved, there
is no observed variable that can block this path.

\textit{Why the formula breaks.}  Stage~1 of the front-door formula uses
$P(M{=}m \mid T{=}t)$ as if the $T \to M$ link were unconfounded.  But
with $U \to M$, the association between $T$ and $M$ is partly due to
their shared cause $U$, not purely the causal effect $T \to M$.
The quantity $P(m \mid t)$ therefore mixes cause and confounding, and
Stage~1 no longer estimates $P(m \mid \doop(T{=}t))$.

\medskip
\noindent\textbf{Case (b): Condition~3 fails --- the $M \to Y$ link has
an extra unobserved confounder.}

Return to the prototypical graph and add a second unobserved variable $V$
that affects both $M$ and $Y$:

\begin{center}
\begin{tikzpicture}[node distance=1.6cm]
  \node[node]  (T) at (0, 0)     {$T$};
  \node[node]  (M) at (2.2, 0)   {$M$};
  \node[node]  (Y) at (4.4, 0)   {$Y$};
  \node[unode] (U) at (1.1,-1.3) {$U$};
  \node[unode] (V) at (3.3,-1.3) {$V$};
  \draw[edge]  (T) -- (M);
  \draw[edge]  (M) -- (Y);
  \draw[dedge] (U) -- (T);
  \draw[dedge] (U) -- (Y);
  \draw[dedge, color=warnred] (V) -- (M)
    node[midway, below left, font=\scriptsize\color{warnred}] {new};
  \draw[dedge, color=warnred] (V) -- (Y)
    node[midway, below right, font=\scriptsize\color{warnred}] {new};
\end{tikzpicture}
\end{center}

\noindent
Conditions~1 and~2 still hold.  But condition~3 now \textbf{fails}: the
path $M \leftarrow V \to Y$ is a back-door path from $M$ to $Y$ that is
\emph{not} blocked by conditioning on $T$.  The path goes
$M \leftarrow V \to Y$ and does not pass through $T$, so fixing $T$ does
nothing to close it.

\textit{Why the formula breaks.}  Stage~2 uses $\sum_{t'} P(y \mid t', m)\,P(t')$
as the back-door-adjusted effect of $M$ on $Y$, with $T$ as the adjustment
variable.  But $T$ only satisfies the back-door criterion for $M \to Y$
when it blocks \emph{all} back-door paths from $M$ to $Y$.  The new path
$M \leftarrow V \to Y$ is not blocked by $T$, so $T$ is no longer a valid
adjustment set for Stage~2, and the Stage~2 quantity is a biased estimate
of $P(y \mid \doop(M{=}m))$.

\medskip
\noindent\textbf{Summary.}
\begin{center}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{@{\hspace{6pt}}lll@{\hspace{6pt}}}
\toprule
\textbf{Modification} & \textbf{Condition violated} & \textbf{Stage broken} \\
\midrule
Add $U \to M$ & Cond.~2: $T \to M$ confounded & Stage~1 \\
Add $V \to M$ and $V \to Y$ & Cond.~3: $M \to Y$ confounded by $V$ & Stage~2 \\
\bottomrule
\end{tabular}
\end{center}
In both cases the criterion detects the failure before any formula is written down: the graphical check on each condition precisely identifies which stage of the two-stage argument collapses.
\end{example}

The prototypical graph where the front-door criterion applies is:
\[
  T \;\to\; M \;\to\; Y, \qquad U \to T, \quad U \to Y,
  \quad \text{no direct } T \to Y \text{ edge.}
\]
The absence of a direct $T \to Y$ edge is not an independent requirement:
it is simply one way of satisfying condition~1, which demands that
\emph{all} directed paths from $T$ to $Y$ pass through $M$.  A direct
$T \to Y$ edge would create a second causal path bypassing $M$, violating
condition~1 and breaking the front-door argument --- as Problem~3(c) below
illustrates.
Here $U$ is an unobserved common cause of $T$ and $Y$; $M$ is a
fully-observed mediator.  Figure~\ref{w3:fig:frontdoor} shows this graph.

\begin{figure}[h]
\centering
\begin{tikzpicture}[node distance=1.8cm]
  \node[node]  (T) at (0,0)      {$T$};
  \node[node]  (M) at (2.0,0)    {$M$};
  \node[node]  (Y) at (4.0,0)    {$Y$};
  \node[unode] (U) at (2.0,-1.4) {$U$};
  \draw[edge]  (T) -- (M);
  \draw[edge]  (M) -- (Y);
  \draw[dedge] (U) -- (T);
  \draw[dedge] (U) -- (Y);
  \node[font=\small\itshape\color{darkgrey}, below=0.15cm of T] {treatment};
  \node[font=\small\itshape\color{darkgrey}, below=0.15cm of M] {mediator};
  \node[font=\small\itshape\color{darkgrey}, below=0.15cm of Y] {outcome};
  \node[font=\small\itshape\color{darkgrey}, below=0.15cm of U] {unobserved};
\end{tikzpicture}
\caption{The prototypical front-door graph.  The confounder $U$ is
unobserved, so back-door adjustment on $U$ is infeasible.  However, the
mediator $M$ is fully observed and satisfies the three front-door
conditions, enabling identification via the front-door formula.}
\label{w3:fig:frontdoor}
\end{figure}

\subsection{The Front-Door Formula}
\label{w3:sec:frontdoor-formula}
\begin{thm}{Front-Door Adjustment Formula \citep{pearl1995causal}}
\label{w3:thm:frontdoor}
If $M$ satisfies the front-door criterion for the effect of $T$ on $Y$, and
$P(t) > 0$ for all $t$, then:
\begin{equation}
  P\!\left(y \mid \doop(T{=}t)\right)
  \;=\; \int
      \underbrace{\Bigl[\int P(y \mid T{=}t',\;M{=}m)\,dP(t')\Bigr]}_{\displaystyle\text{Stage 2}}
      \;\underbrace{dP(m \mid t)}_{\displaystyle\text{Stage 1}}.
  \label{w3:eq:frontdoor}
\end{equation}
Equivalently, writing $\mu(t',m) = \E[Y \mid T{=}t',\;M{=}m]$,
\begin{equation}
  \E\!\left[Y \mid \doop(T{=}t)\right]
  \;=\; \E_{M \mid T=t}\!\Bigl[\E_T\bigl[\mu(T,\,M)\bigr]\Bigr].
  \label{w3:eq:frontdoor-exp}
\end{equation}
\textbf{Stage~1}: $dP(m \mid t)$ is the distribution of the mediator $M$
given $T{=}t$.  Front-door condition~2 guarantees no confounding between
$T$ and $M$, so this conditional distribution is directly identifiable
from the observed data.\\[4pt]
\textbf{Stage~2}: $\int P(y \mid t',\,m)\,dP(t')$ is the back-door-adjusted
causal effect of $M$ on $Y$, with $T$ serving as the adjustment variable.
In the iterated-expectation form~\eqref{w3:eq:frontdoor-exp}, this is
$\E_T[\mu(T,M)]$: the regression function $\mu(T,M)$ averaged over the
\emph{marginal} distribution of $T$, not the conditional distribution
$P(T \mid M{=}m)$.  Front-door condition~3 guarantees that conditioning
on $T$ blocks all back-door paths from $M$ to $Y$ (specifically,
$M \leftarrow T \leftarrow U \to Y$), so this marginalization removes the
confounding introduced by $U$.\\[4pt]
The formula chains the two stages by integrating Stage~2 against the
Stage~1 measure $dP(m \mid t)$.  When $T$ and $M$ are both discrete,
the integrals reduce to the summation form
$\sum_m P(m\mid t)\sum_{t'} P(y\mid t',m)\,P(t')$.
\end{thm}

\noindent\textit{Derivation sketch.}
The formula is the composition of two back-door applications.
\begin{enumerate}
  \item \textbf{Identify $P(m \mid \doop(T{=}t))$.}
        By condition~2 there are no back-door paths from $T$ to $M$,
        so $T$ satisfies the (trivial) back-door criterion for its effect
        on $M$ with empty adjustment set.  Hence
        $P(m \mid \doop(T{=}t)) = P(m \mid T{=}t)$.

  \item \textbf{Identify $P(y \mid \doop(M{=}m))$.}
        By condition~3, every back-door path from $M$ to $Y$ passes
        through $T$.  Conditioning on $T$ blocks those paths, so $T$ is
        a valid back-door adjustment set for the $M \to Y$ effect:
        \[
          P(y \mid \doop(M{=}m))
          \;=\; \int P(y \mid T{=}t',\;M{=}m)\,dP(t').
        \]

  \item \textbf{Chain the two stages.}
        By condition~1, every directed path from $T$ to $Y$ passes
        through $M$, so the effect of $T$ on $Y$ operates entirely
        \emph{through} $M$.  Combining steps~1 and~2:
        \[
          P(y \mid \doop(T{=}t))
          = \int P(y \mid \doop(M{=}m))\,dP(m \mid \doop(T{=}t))
          = \int \Bigl[\int P(y \mid t',m)\,dP(t')\Bigr]\,dP(m \mid t).
        \]
\end{enumerate}
The role of each condition is now transparent: condition~1 licenses
step~3 (the chaining), condition~2 licenses step~1 (replacing
$\doop(T{=}t)$ with conditioning), and condition~3 licenses step~2
(back-door adjustment for $M \to Y$).

A formal proof via three applications of the do-calculus rules is given in
Section~\ref{w3:sec:proofs}.

\begin{figure}[t]
\centering
\begin{tikzpicture}[node distance=1.8cm and 2.0cm]
  \node[node]  (T) at (0,0)     {$T$};
  \node[node]  (M) at (3.2,0)   {$M$};
  \node[node]  (Y) at (6.4,0)   {$Y$};
  \node[unode] (U) at (3.2,2.4) {$U$};

  \draw[edge]  (T) -- (M);
  \draw[edge]  (M) -- (Y);
  \draw[dedge] (U) -- (T);
  \draw[dedge] (U) -- (Y);

  \node[font=\small\itshape\color{darkgrey}, align=center]
    at (1.6,-0.90) {Stage~1\\identify $T \to M$};
  \node[font=\small\itshape\color{darkgrey}, align=center]
    at (4.8,-0.90) {Stage~2\\identify $M \to Y \mid T$};

  \node[font=\small\itshape\color{darkgrey}]
    at (3.2,3.10) {$U$ unobserved};

  \node[font=\scriptsize\itshape\color{darkgrey}, align=center]
    at (3.2,-1.85)
    {(conditioning on $T$ blocks $M \leftarrow T \leftarrow U \to Y$,
      making the $M \to Y$ effect identifiable)};
\end{tikzpicture}
\caption{Why the front-door criterion works.  The total effect of $T$ on $Y$
is not directly identifiable because $U$ confounds $T$ and $Y$.  However,
the effect of $T$ on the mediator $M$ is unconfounded (Stage~1), and the
effect of $M$ on $Y$ becomes identifiable after conditioning on $T$, which
blocks the back-door path through $U$ (Stage~2).  The front-door formula
reconstructs the total causal effect by combining these two identifiable
pieces.}
\label{w3:fig:frontdoor-two-stage}
\end{figure}

\subsection{Interpretation: Two-Stage Causal Accounting}
\label{w3:sec:twostage}
The two-stage structure annotated in equation~\eqref{w3:eq:frontdoor} has a
natural graphical reading in Figure~\ref{w3:fig:frontdoor-two-stage}: Stage~1
traverses the unconfounded $T \to M$ link, and Stage~2 traverses the
$T$-adjusted $M \to Y$ link.  Together they reconstruct the total causal
effect $T \to Y$ without ever directly comparing the treatment groups on
$Y$, which is where the confounding by $U$ would otherwise enter.

\begin{example}{Smoking, Tar, and Cancer}
\label{w3:ex:smoking}
The classic front-door example \citep{pearl1995causal} is:
$T$ = smoking, $M$ = tar deposits in lungs, $Y$ = lung cancer,
$U$ = genetic predisposition (unobserved).
The graph is $T \to M \to Y$ with $U \to T$ and $U \to Y$.
There is no direct $T\to Y$ path (all effect of smoking operates through
tar).

Conditions~1--3 are satisfied: all causal paths from $T$ to $Y$ pass
through $M$; no back-door path exists from $T$ to $M$ (there is no arrow
from $U$ into $M$); and the only back-door path from $M$ to $Y$ is
$M \leftarrow T \leftarrow U \to Y$, which is blocked by conditioning on $T$.

The front-door formula gives:
\[
  P(\text{cancer} \mid \doop(T{=}\text{smoke}))
  = \int\!\left[\int P(\text{cancer} \mid t', m)\,dP(t')\right]
    dP(m \mid \text{smoke}),
\]
which is entirely computable from the observed joint distribution of
$(T, M, Y)$, without any data on $U$.  For the discrete case this becomes
a double sum over $(m, t')$.
\end{example}

\begin{example}{Front-Door Formula: Binary Numerical Computation}
\label{w3:ex:numerical}
Let $T, M, Y \in \{0, 1\}$ with $U$ unobserved and the graph
$T \to M \to Y$, $U \to T$, $U \to Y$.  The observed joint distribution
$P(T, M, Y)$ is given in the following population table.

\medskip
\begin{center}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{@{\hspace{8pt}}ccc@{\hspace{16pt}}c@{\hspace{8pt}}}
\toprule
$T$ & $M$ & $Y$ & $P(T,M,Y)$ \\
\midrule
0 & 0 & 0 & 0.360 \\
0 & 0 & 1 & 0.040 \\
0 & 1 & 0 & 0.050 \\
0 & 1 & 1 & 0.050 \\
1 & 0 & 0 & 0.070 \\
1 & 0 & 1 & 0.030 \\
1 & 1 & 0 & 0.120 \\
1 & 1 & 1 & 0.280 \\
\bottomrule
\end{tabular}
\end{center}

\medskip\noindent
All quantities needed for the front-door formula can be read off from this
table by marginalization.  For example,
$P(T{=}1) = 0.070 + 0.030 + 0.120 + 0.280 = 0.5$,
and
$P(M{=}1 \mid T{=}1) = (0.120 + 0.280)/(0.5) = 0.8$.

\medskip\noindent
We compute $P(Y{=}1 \mid \doop(T{=}t))$ for $t \in \{0,1\}$ using the
front-door formula.  The inner sum (Stage~2) does not depend on the
intervention value $t$, so compute it once for each value of $M$:
\[
  S_0 \;=\; \sum_{t'} P(Y{=}1 \mid t', M{=}0)\,P(t')
         = 0.1 \times 0.5 + 0.3 \times 0.5 = 0.20,
\]
\[
  S_1 \;=\; \sum_{t'} P(Y{=}1 \mid t', M{=}1)\,P(t')
         = 0.5 \times 0.5 + 0.7 \times 0.5 = 0.60.
\]
Now apply the outer sum (Stage~1) for each intervention value:
\begin{align*}
  P(Y{=}1 \mid \doop(T{=}1))
    &= P(M{=}0 \mid T{=}1)\,S_0 + P(M{=}1 \mid T{=}1)\,S_1 \\
    &= 0.2 \times 0.20 \;+\; 0.8 \times 0.60 \;=\; 0.52, \\[4pt]
  P(Y{=}1 \mid \doop(T{=}0))
    &= P(M{=}0 \mid T{=}0)\,S_0 + P(M{=}1 \mid T{=}0)\,S_1 \\
    &= 0.8 \times 0.20 \;+\; 0.2 \times 0.60 \;=\; 0.28.
\end{align*}
The average causal effect is $0.52 - 0.28 = 0.24$.

\paragraph{Comparison with the naive approach.}
A naive analyst ignoring the unobserved confounder $U$ would estimate the
effect of $T$ on $Y$ by the observational contrast
$P(Y{=}1 \mid T{=}1) - P(Y{=}1 \mid T{=}0)$, read directly from the
population table:
\begin{align*}
  P(Y{=}1 \mid T{=}1)
    &= \frac{0.030 + 0.280}{0.5} = 0.62, \\[4pt]
  P(Y{=}1 \mid T{=}0)
    &= \frac{0.040 + 0.050}{0.5} = 0.18.
\end{align*}
The naive estimate of the causal effect is $0.62 - 0.18 = 0.44$, which is
\emph{almost twice} the true causal effect of $0.24$.  The bias arises
because $U$ simultaneously increases $T$ and $Y$: high-$U$ individuals
tend to have $T{=}1$ and $Y{=}1$ for reasons unrelated to the $T\to M\to Y$
pathway, inflating the observational association.  The front-door formula
removes this bias by routing the effect estimate through the unconfounded
$T\to M$ link and the $T$-adjusted $M\to Y$ link, never directly comparing
the $T{=}1$ and $T{=}0$ groups on $Y$.

Note how Stage~2 (the inner sum $S_m$) averages $P(Y\mid t', m)$ over the
\emph{marginal} distribution $P(t')$, not the treatment-conditional
distribution $P(t' \mid m)$.  This re-weighting is precisely what removes
the confounding introduced by $U$: the Stage~2 quantity $S_m$ is the
back-door-adjusted effect of $M$ on $Y$, with $T$ serving as the
adjustment variable.
\end{example}

\begin{warn}{Comparing Back-Door and Front-Door}
The two criteria are \emph{not} nested.  Neither implies the other.
\begin{itemize}
  \item The back-door criterion requires all confounders to be observed
        (or blocked by observed variables), but places no restriction on
        mediators.
  \item The front-door criterion requires an observed mediator intercepting
        all causal paths, but the confounder $U$ can be entirely
        unobserved.
\end{itemize}
There exist graphs where neither criterion alone applies, but where the
full do-calculus (or the ID algorithm) can still identify the causal
effect.
\end{warn}

\medskip\noindent
Up to this point, the chapter has focused on two concrete identification
criteria that can be checked directly in common graphical settings.
We now turn to the more general algebraic machinery that subsumes both.
The purpose of the do-calculus is not to replace back-door and front-door
adjustment with something unrelated, but to show that both are instances
of a single system for transforming interventional expressions into
observational ones.

% ==============================================================
\section{The Three Rules of Do-Calculus}
\label{w3:sec:rules}
% ==============================================================

\noindent\textbf{General theory.}\enspace
The remainder of this chapter develops the general calculus of
identification and the limits of what can be identified from observational
data alone.

\medskip
Having seen two concrete identification strategies, we now develop the
general algebraic machinery that underlies both.  \citet{pearl1995causal}
introduced the three rules of do-calculus as a system for manipulating
expressions involving the do-operator: each rule licenses a transformation
that removes or exchanges terms, bringing the expression one step closer
to a purely observational formula.  The back-door and front-door formulas
of Sections~\ref{w3:sec:backdoor} and~\ref{w3:sec:frontdoor} are each
derivable as three-step sequences of these rules, as shown in
Section~\ref{w3:sec:proofs}.  Whether the three rules are
\emph{sufficient} to identify every identifiable causal effect is a deeper
question; Section~\ref{w3:sec:completeness} presents the completeness
theorem of \citet{shpitser2006identification}, which answers it
affirmatively.

\begin{remark}
\textbf{First reading / second reading.}
On a first reading, focus on the qualitative role of the three rules:
Rule~1 adds or removes observations, Rule~2 exchanges an action for an
observation, and Rule~3 adds or removes actions.
The detailed graph subscripts and formal statements can be absorbed
gradually.
The main conceptual point is that each rule tests whether a variable
becomes irrelevant after the appropriate arrows have been deleted from
the graph --- exactly the same kind of d-separation check introduced in
Chapter~2.
On a second reading, work through the graphical conditions in the
statement table and verify them on the examples in
Section~\ref{w3:sec:rules-intuition}.
\end{remark}

\subsection{Statement of the Rules}
\label{w3:sec:rules-statement}
Recall from Section~\ref{w2:sec:markov} of Chapter~2 that a distribution
$P$ is \emph{Markov with respect to $\Gcal$} if every d-separation statement in $\Gcal$
implies the corresponding conditional independence in $P$:
\[
  (X \indep Y \mid \mathbf{S})_{\Gcal}
  \;\Longrightarrow\;
  X \indep Y \mid \mathbf{S} \text{ under } P.
\]
Equivalently, $P$ admits the Markov factorization
$P(v_1,\ldots,v_n) = \prod_i P(v_i \mid \Pa(v_i))$.

\paragraph{How to read the graphical conditions in Theorem~\ref{w3:thm:docalculus}.}
Each rule has the form \emph{``expression~A = expression~B if
$(Y \indep Z \mid X, W)_{\Gcal'}$''}, where $\Gcal'$ is a specific
intervention graph built from $\Gcal$ by two deletions.
The subscript encodes both deletions in compressed notation.
The graph subscripts may look heavier than the ideas themselves.
They are simply bookkeeping devices that record which incoming or
outgoing arrows were deleted before the d-separation condition is checked.
The table below gives the recipe for each subscript that appears.

\medskip
\begin{center}
\renewcommand{\arraystretch}{1.45}
\begin{tabular}{@{\hspace{6pt}}l@{\hspace{14pt}}l@{\hspace{14pt}}l@{\hspace{6pt}}}
\toprule
\textbf{Subscript} & \textbf{Step 1: delete} & \textbf{Step 2: delete} \\
\midrule
$\Gcal_{\overline{X}\,\underline{Z}}$
  & all arrows \emph{into} $X$
  & all arrows \emph{out of} $Z$ \\
$\Gcal_{\overline{X}\,\overline{Z(W)}}$
  & all arrows \emph{into} $X$
  & arrows \emph{into} $Z$ except those from $W$ \\
$\Gcal_{\overline{X}\,\overline{Z(\overline{W})}}$
  & all arrows \emph{into} $X$
  & arrows \emph{into} $Z$ that are not from ancestors of $W$
    in $\Gcal_{\overline{X}}$ \\
\bottomrule
\end{tabular}
\end{center}

\medskip\noindent
Having formed $\Gcal'$ by these two deletions, the condition
$(Y \indep Z \mid X, W)_{\Gcal'}$ is then checked by the same
d-separation algorithm used in Chapter~2: find all paths between $Y$ and
$Z$, and verify that every path is blocked by $\{X, W\}$ in $\Gcal'$.

\medskip\noindent
\textbf{The essential pattern.}
At first reading, the notation of the do-calculus can look more
complicated than the ideas themselves.  The underlying logic is always
the same: each rule asks whether a certain observation or intervention
is \emph{irrelevant} once the appropriate arrows have been removed from
the graph.  Remove the arrows, check d-separation, and if the check
passes, the rule applies.  The graph subscripts are
\emph{bookkeeping devices} that record exactly which arrows were deleted
before the d-separation condition is checked --- nothing more.
Once that is clear, the three rules reduce to three instances of a
single question: \emph{``In the simplified graph obtained by this
particular surgery, does the relevant d-separation hold?''}

\begin{thm}{The Three Rules of Do-Calculus \citep{pearl1995causal}}
\label{w3:thm:docalculus}
Let $\Gcal$ be a DAG over variables $\mathcal{V}$, and let $X, Y, Z, W
\subseteq \mathcal{V}$ be disjoint sets.  The rules hold for any
distribution $P$ Markov with respect to $\Gcal$.

\medskip
\noindent\textbf{Rule 1 (Insertion/Deletion of Observations).}
\[
  P(y \mid \doop(x), z, w) \;=\; P(y \mid \doop(x), w)
  \quad\text{if}\quad
  (Y \indep Z \mid X, W)_{\Gcal_{\overline{X}\,\underline{Z}}}.
\]
If $Y$ and $Z$ are d-separated in the graph with arrows into $X$ deleted
\emph{and} arrows out of $Z$ deleted, then conditioning on $Z$ has no
effect on the distribution of $Y$ (given $\doop(x)$ and $W$).

\smallskip
\noindent\textit{Instance} ($Z \to T \to Y$, intervene on $T$).
Set $X{=}T$, $Z{=}Z$, $W{=}\varnothing$.
Form $\Gcal_{\overline{T}\,\underline{Z}}$: delete $Z\to T$ (arrow into $T$),
then delete all outgoing arrows of $Z$ — but $Z \to T$ was already gone, so $Z$
becomes isolated.  Since $(Y \indep Z)_{\Gcal_{\overline{T}\,\underline{Z}}}$
holds, Rule~1 gives $P(y \mid \doop(t), z) = P(y \mid \doop(t))$: knowing the
former cause $Z$ of $T$ is irrelevant once $T$ has been set externally.

\medskip
\noindent\textbf{Rule 2 (Action/Observation Exchange).}
\[
  P(y \mid \doop(x), \doop(z), w) \;=\; P(y \mid \doop(x), z, w)
  \quad\text{if}\quad
  (Y \indep Z \mid X, W)_{\Gcal_{\overline{X}\,\overline{Z(W)}}}.
\]
When the condition holds, intervening on $Z$ is equivalent to conditioning
on $Z$ (within the scope of $\doop(x)$ and $W$).

\smallskip
\noindent\textit{Instance} ($Z \to T \to Y$, condition on $T$).
Set $X{=}\varnothing$, $Z{=}Z$, $W{=}\{T\}$.
Form $\Gcal_{\overline{Z(T)}}$: delete arrows into $Z$ except those from
$T$ — but $Z$ has no parents, so $\Gcal_{\overline{Z(T)}} = \Gcal$.
The only path from $Z$ to $Y$ is $Z \to T \to Y$, which is a chain
blocked by $T \in W$.  Since $(Y \indep Z \mid T)_{\Gcal_{\overline{Z(T)}}}$
holds, Rule~2 gives $P(y \mid \doop(z), t) = P(y \mid z, t)$: with $T$
observed, intervening on $Z$ versus merely observing $Z{=}z$ makes no
difference to $Y$.

\medskip
\noindent\textbf{Rule 3 (Insertion/Deletion of Actions).}
\[
  P(y \mid \doop(x), \doop(z), w) \;=\; P(y \mid \doop(x), w)
  \quad\text{if}\quad
  (Y \indep Z \mid X, W)_{\Gcal_{\overline{X}\,\overline{Z(\overline{W})}}}.
\]
When the condition holds, the intervention $\doop(z)$ can be deleted
entirely.

\smallskip
\noindent\textit{Instance} ($Z \to T \to Y$, already intervening on $T$).
Set $X{=}T$, $Z{=}Z$, $W{=}\varnothing$.
Form $\Gcal_{\overline{T}\,\overline{Z(\overline{\varnothing})}} =
\Gcal_{\overline{T}\,\overline{Z}}$: delete arrows into $T$ (removes
$Z \to T$), then delete arrows into $Z$ (none exist).  Node $Z$ is
now isolated with no path to $Y$, so
$(Y \indep Z)_{\Gcal_{\overline{T}\,\overline{Z}}}$ holds.  Rule~3 gives
$P(y \mid \doop(t), \doop(z)) = P(y \mid \doop(t))$: with $T$ already
fixed, setting $Z$ to any value cannot reach $Y$, so $\doop(z)$ is
superfluous.
\end{thm}

\begin{remark}
Each rule is a \emph{conditional independence statement in a specific
intervention graph}.  The three rules cover all ways of adding or removing
terms from a do-expression: Rule~1 handles observations; Rule~2 exchanges
actions for observations; Rule~3 deletes actions.  Any sequence of such
operations that converts a do-expression into a purely observational
expression is an \emph{identification proof}.
\end{remark}

\begin{remark}[Soundness of the Rules]
\label{w3:rem:soundness}
The three rules are \emph{sound}: each holds for every distribution $P$
Markov with respect to $\Gcal$.  The argument for Rule~2 is instructive.
Recall that the post-intervention distribution $P(\cdot \mid \doop(x))$
is Markov with respect to $\Gcal_{\overline{X}}$: the Markov factorization
of $\Gcal_{\overline{X}}$ is exactly $P$ with the structural equation
for $X$ replaced by the constant $x$ (the truncated factorization).
Rule~2 asserts $P(y \mid \doop(x), \doop(z), w) = P(y \mid \doop(x), z, w)$
when $(Y \indep Z \mid X, W)_{\Gcal_{\overline{X}\,\overline{Z(W)}}}$.
\emph{Why does this hold?}  In the post-$\doop(x)$ world, the
distribution is Markov with respect to $\Gcal_{\overline{X}}$.  Further
intervening on $Z$ gives a distribution Markov with respect to
$\Gcal_{\overline{X}\,\overline{Z}}$, a sub-graph of
$\Gcal_{\overline{X}\,\overline{Z(W)}}$.  The d-separation condition
$(Y \indep Z \mid X, W)_{\Gcal_{\overline{X}\,\overline{Z(W)}}}$ therefore
implies $Y \indep Z \mid W$ in the post-$\doop(x,z)$ distribution.  Since
the distribution of $Y$ is the same whether we \emph{intervene} on $Z$ or
\emph{observe} $Z$ — precisely because the d-separation blocks all
back-door paths from $Z$ to $Y$ that $\doop(z)$ would additionally sever
— the two expressions are equal.  Rules~1 and~3 are analogous: Rule~1
checks that observing $Z$ is irrelevant (d-separation in
$\Gcal_{\overline{X}\,\underline{Z}}$); Rule~3 checks that intervening on
$Z$ has no downstream effect on $Y$ (d-separation in
$\Gcal_{\overline{X}\,\overline{Z(\overline{W})}}$).
Full proofs appear in \citet{pearl1995causal}.
\end{remark}

\begin{center}
\renewcommand{\arraystretch}{1.35}
\begin{tabular}{@{\hspace{6pt}}c@{\hspace{12pt}}|@{\hspace{12pt}}l@{\hspace{6pt}}}
\textbf{Rule} & \textbf{Interpretation} \\
\hline
Rule 1 & Insert or delete \emph{observations} from a conditioning set \\
Rule 2 & Exchange an \emph{intervention} $\doop(z)$ for an \emph{observation} $z$ \\
Rule 3 & Insert or delete \emph{interventions} from an expression \\
\end{tabular}
\end{center}

\subsection{Intuition for Each Rule}
\label{w3:sec:rules-intuition}
\paragraph{Rule 1 --- Adding/Removing Observations.}
The condition $(Y \indep Z \mid X, W)_{\Gcal_{\overline{X}\,\underline{Z}}}$
says: in the post-$\doop(x)$ world, if we further deactivate $Z$'s causal
influence (delete its outgoing arrows), $Y$ and $Z$ become independent given
$W$.  This means $Z$ carries no information about $Y$ that is not already
in $W$, so it can be freely inserted or removed from the conditioning set.

\begin{example}{Rule 1 --- Deleting a Redundant Observation}
\label{w3:ex:rule1}
Graph: $Z \to T \to Y$, no other edges.  Suppose we intervene on $T$.

\begin{center}
\begin{tikzpicture}
  \node[node] (Z) at (0,0)   {$Z$};
  \node[node] (T) at (1.8,0) {$T$};
  \node[node] (Y) at (3.6,0) {$Y$};
  \draw[edge] (Z) -- (T);
  \draw[edge] (T) -- (Y);
\end{tikzpicture}
\end{center}

\noindent
In $\Gcal_{\overline{T}\,\underline{Z}}$: delete arrows into $T$ (removes
$Z \to T$), then delete arrows out of $Z$ (nothing remains).  Node $Z$ is
now isolated, so $(Y \indep Z \mid T)$ holds trivially.  Rule~1 gives:
\[
  P(y \mid \doop(t), z) \;=\; P(y \mid \doop(t)).
\]
\textit{Interpretation}: the intervention $\doop(t)$ severs $Z$'s only
route to $Y$.  Once we fix $T$ externally, knowing $Z$ (the former cause of
$T$) carries no additional information about $Y$.
\end{example}

\paragraph{Rule 2 --- Swapping Action for Observation.}
This is the most frequently used rule.  The condition
$(Y \indep Z \mid X, W)_{\Gcal_{\overline{X}\,\overline{Z(W)}}}$ says: in
a graph where we delete all back-door paths into $Z$ (except those from
$W$), $Y$ and $Z$ are d-separated given $X$ and $W$.  Intuitively, if
the only paths from $Z$ to $Y$ are causal (forward) paths, then intervening
on $Z$ and conditioning on $Z$ have the same effect on $Y$ --- there is no
confounding between $Z$ and $Y$.

\begin{example}{Rule 2 --- Action/Observation Exchange}
\label{w3:ex:rule2}
Graph: $X \to Z \to Y$, no other edges.  Suppose we have already intervened
on $X$ and wish to exchange the intervention $\doop(z)$ for the observation
$z$ in $P(y \mid \doop(x), \doop(z))$.

\medskip
\noindent\textbf{Original graph.}
\begin{center}
\begin{tikzpicture}
  \node[node] (X) at (0,0)   {$X$};
  \node[node] (Z) at (1.8,0) {$Z$};
  \node[node] (Y) at (3.6,0) {$Y$};
  \draw[edge] (X) -- (Z);
  \draw[edge] (Z) -- (Y);
\end{tikzpicture}
\end{center}

\noindent\textbf{Modified graph} $\Gcal_{\overline{X}\,\overline{Z}}$:
delete arrows into $X$ (none to remove) and all arrows into $Z$ (removes
$X \to Z$).  Node $Z$ has no incoming arrows in the modified graph.

\begin{center}
\begin{tikzpicture}
  \node[node] (X) at (0,0)   {$X$};
  \node[node] (Z) at (1.8,0) {$Z$};
  \node[node] (Y) at (3.6,0) {$Y$};
  \draw[xedge] (X) -- (Z);
  \draw[edge]  (Z) -- (Y);
\end{tikzpicture}
\end{center}

\noindent\textbf{D-separation check}: in the modified graph, $Z$ has
no incoming arrows, so there are no back-door paths from $Z$ to $Y$.
The only path is the causal $Z \to Y$.  Since $Y$ depends on $Z$ only
through $Z$'s value (not through the mechanism that produced $Z$),
$(Y \indep Z)_{\Gcal_{\overline{X}\,\overline{Z}}}$ holds in the sense
that no spurious association between $Z$ and $Y$ can arise. \checkmark

\noindent\textbf{Rule~2 gives}:
\[
  P(y \mid \doop(x), \doop(z)) \;=\; P(y \mid \doop(x), z).
\]
\textit{Interpretation}: $Y$ depends on $Z$ only through $Z$'s value, not
through the causal mechanism that generated $Z$.  Externally fixing $Z$
and merely observing $Z{=}z$ are therefore equivalent for $Y$: in both
cases, $Z$'s downstream effect on $Y$ is exactly the same.
\end{example}

\paragraph{Rule 3 --- Deleting an Intervention.}
The condition says that $Z$ has no causal path to $Y$ that is not blocked by
$X$ or $W$ in the appropriate intervention graph.  Under this condition,
setting $Z$ to any value has no effect on $Y$, so $\doop(z)$ can be dropped.

\begin{example}{Rule 3 --- Deleting an Irrelevant Intervention}
\label{w3:ex:rule3}
Graph: $Z \to X \to Y$, no other edges.  Suppose we intervene on both $X$
and $Z$, and wish to drop $\doop(z)$ from $P(y \mid \doop(x), \doop(z))$.

\medskip
\noindent\textbf{Original graph.}
\begin{center}
\begin{tikzpicture}
  \node[node] (Z) at (0,0)   {$Z$};
  \node[node] (X) at (1.8,0) {$X$};
  \node[node] (Y) at (3.6,0) {$Y$};
  \draw[edge] (Z) -- (X);
  \draw[edge] (X) -- (Y);
\end{tikzpicture}
\end{center}

\noindent\textbf{Modified graph} $\Gcal_{\overline{X}\,\overline{Z}}$:
delete all arrows into $X$ (removes $Z \to X$), then delete all arrows
into $Z$ (none to remove).  Node $Z$ is now isolated.

\begin{center}
\begin{tikzpicture}
  \node[node] (Z) at (0,0)   {$Z$};
  \node[node] (X) at (1.8,0) {$X$};
  \node[node] (Y) at (3.6,0) {$Y$};
  \draw[xedge] (Z) -- (X);
  \draw[edge]  (X) -- (Y);
\end{tikzpicture}
\end{center}

\noindent\textbf{D-separation check}: in the modified graph, $Z$ is
isolated.  There is no path from $Z$ to $Y$, so
$(Y \indep Z)_{\Gcal_{\overline{X}\,\overline{Z}}}$ holds trivially.
\checkmark

\noindent\textbf{Rule~3 gives}:
\[
  P(y \mid \doop(x), \doop(z)) \;=\; P(y \mid \doop(x)).
\]
\textit{Interpretation}: once $X$ is fixed externally, $Z$'s only route
to $Y$ (through $X$) is severed.  Setting $Z$ to any value cannot change
$Y$, so the intervention $\doop(z)$ is superfluous and can be deleted.

\textit{Contrast with Rule~1}: if instead of intervening on $Z$ we merely
\emph{observe} it, Example~\ref{w3:ex:rule1} shows that the same
d-separation argument lets us drop the \emph{observation} $z$ from
$P(y \mid \doop(x), z)$.  Rule~1 deletes a redundant observation;
Rule~3 deletes a redundant intervention.
\end{example}

\begin{example}{A Two-Rule Derivation: the Proxy Variable Graph}
\label{w3:ex:tworule}
The three preceding examples each applied one rule to a simple chain.
We now chain two rules in sequence to derive a non-trivial identification
formula for a graph that requires both.

\medskip
\noindent\textbf{Graph.}
$T$ has a direct effect on $Y$.  An unobserved variable $U$ is a common
cause of $T$ and an observed proxy $S$; $S$ in turn causes $Y$.
$U$ does \emph{not} directly cause $Y$.

\begin{center}
\begin{tikzpicture}[node distance=1.8cm]
  \node[node]  (T) at (0, 0)    {$T$};
  \node[unode] (U) at (1.6, 1.2){$U$};
  \node[node]  (S) at (3.2, 1.2){$S$};
  \node[node]  (Y) at (3.2, 0)  {$Y$};
  \draw[edge]  (T) -- (Y);
  \draw[dedge] (U) -- (T);
  \draw[dedge] (U) -- (S);
  \draw[edge]  (S) -- (Y);
\end{tikzpicture}
\end{center}

\noindent
Because $U$ is unobserved, back-door adjustment directly on $U$ is
infeasible.  However, $S$ lies on the back-door path $T \leftarrow U \to
S \to Y$, and $S$ is observed.  We now derive the identification formula
using the rules, making every graphical check explicit.

\medskip
\noindent\textbf{Goal}: simplify $P(y \mid \doop(t))$ to a purely
observational expression.

\medskip
\noindent\textbf{Step 1.  Introduce $S$ by the law of total probability.}
\[
  P(y \mid \doop(t))
  \;=\; \sum_s P(y \mid \doop(t),\,s)\;P(s \mid \doop(t)).
\]
This produces two sub-expressions: the conditional
$P(y \mid \doop(t), s)$ and the marginal $P(s \mid \doop(t))$.
Each is handled by a separate rule.

\medskip
\noindent\textbf{Step 2.  Rule~3 applied to $P(s \mid \doop(t))$.}

\emph{Claim}: $P(s \mid \doop(t)) = P(s)$.

\emph{Rule~3 condition} (with $X{=}\varnothing$, $Z{=}T$, $W{=}\varnothing$):
$(S \indep T)_{\Gcal_{\overline{T}}}$.

\emph{Graphical check}: form $\Gcal_{\overline{T}}$ by deleting the arrow
$U \to T$.  The remaining edges are $T \to Y$, $U \to S$, $S \to Y$.
In this graph, $T$ has no parents and no edge connects $T$ to $S$ or to
the $U$--$S$ component.  There is no path between $T$ and $S$, so
$(S \indep T)_{\Gcal_{\overline{T}}}$. \checkmark

\emph{Conclusion}: $P(s \mid \doop(t)) = P(s)$.

\medskip
\noindent\textbf{Step 3.  Back-door d-separation argument applied to
$P(y \mid \doop(t), s)$.}

\emph{Claim}: $P(y \mid \doop(t), s) = P(y \mid t, s)$.

\emph{Condition} (back-door criterion with $\mathbf{S}=\{S\}$):
$(Y \indep T \mid S)_{\Gcal_{\underline{T}}}$.

\emph{Graphical check}: form $\Gcal_{\underline{T}}$ by deleting the
arrow $T \to Y$.  The remaining edges are $U \to T$, $U \to S$, $S \to Y$.
Paths between $T$ and $Y$ in $\Gcal_{\underline{T}}$: only the back-door
path $T \leftarrow U \to S \to Y$.  Node $S$ is a non-collider on this
path; conditioning on $S$ blocks it. \checkmark

\emph{Conclusion}: by the truncated factorization, $\{S\}$ satisfies the
back-door criterion for $(T, Y)$, so $P(y \mid \doop(t), s) = P(y \mid t, s)$.

\medskip
\noindent\textbf{Combining Steps 1--3:}
\[
  P(y \mid \doop(t))
  \;=\; \sum_s P(y \mid t,\,s)\;P(s).
\]
The causal effect of $T$ on $Y$ is identified by standardizing the
conditional distribution $P(y \mid t, s)$ over the population marginal
$P(s)$ --- even though the confounder $U$ is entirely unobserved.

\medskip
\noindent\textbf{What each rule did.}
Step~2 used Rule~3 to \emph{decouple $S$ from the intervention}: the
d-separation check confirmed that $T$ has no causal route to $S$ in the
post-intervention world, so $P(s \mid \doop(t)) = P(s)$.
Step~3 used the back-door d-separation in $\Gcal_{\underline{T}}$ to
\emph{justify treating the intervention $\doop(t)$ as an observation $t$}
within each stratum of $S$.  Neither step alone suffices: without
Step~2, the marginal retains a $\doop(t)$; without Step~3, the
conditional cannot be converted to an observational quantity.
The complete formula follows only after both rules are applied to
their respective sub-expressions.
\end{example}

\subsection{Proofs of the Identification Formulas}
\label{w3:sec:proofs}
With the three rules now in hand, we can prove the two identification
formulas stated in Sections~\ref{w3:sec:backdoor}
and~\ref{w3:sec:frontdoor}.

\begin{tcolorbox}[defstyle, title={Proof of Theorem~\ref{w3:thm:backdoor} (Back-Door Formula) via Rule~3 and the Back-Door d-Separation Argument}]

\noindent\textit{Graph structure}: $T \to Y$, unobserved $U \to T$,
$U \to Y$, and observed $\mathbf{S}$ blocking the back-door path
$T \leftarrow U \to Y$.

\begin{center}
\begin{tikzpicture}[node distance=1.6cm]
  \node[node]  (T) at (0, 0)     {$T$};
  \node[node]  (Y) at (3.0, 0)   {$Y$};
  \node[unode] (U) at (1.5, 1.2) {$U$};
  \node[node]  (Z) at (1.5,-1.2) {$\mathbf{S}$};
  \draw[edge]  (T) -- (Y);
  \draw[dedge] (U) -- (T);
  \draw[dedge] (U) -- (Y);
  \draw[edge]  (Z) -- (T);
  \draw[edge]  (Z) -- (Y);
  \node[font=\scriptsize\color{warnred}] at (1.5, 0.35)
    {\textit{back-door path}};
\end{tikzpicture}
\end{center}

\noindent The back-door path $T \leftarrow U \to Y$ is shown dashed.
$\mathbf{S}$ blocks this path.  The proof manipulates $P(y \mid \doop(t))$
in three steps.

\medskip
\noindent\textbf{Step 1: Introduce $\mathbf{S}$ by the law of total
probability.}
\[
  P(y \mid \doop(t))
  = \int P(y \mid \doop(t),\;\mathbf{s})\,dP(\mathbf{s} \mid \doop(t)).
\]

\noindent\textbf{Step 2: Simplify $P(\mathbf{s} \mid \doop(t))$.}
Because $\mathbf{S}$ contains no descendants of $T$ (condition~1),
intervening on $T$ cannot causally affect any node in $\mathbf{S}$.
The graphical condition for Rule~3 is $(\mathbf{S} \indep T)_{\Gcal_{\overline{T}}}$:
in $\Gcal_{\overline{T}}$, all arrows into $T$ are deleted so $T$ has no
parents, and no directed path from $T$ reaches any node in $\mathbf{S}$
(by condition~1).  Therefore $(\mathbf{S} \indep T)_{\Gcal_{\overline{T}}}$
holds, and Rule~3 gives:
\[
  P(\mathbf{s} \mid \doop(t)) = P(\mathbf{s}).
\]

\noindent\textbf{Step 3: Convert $P(y \mid \doop(t), \mathbf{s})$ to an
observation.}
Condition~2 ensures that, within each stratum $\mathbf{S}=\mathbf{s}$,
all back-door paths from $T$ to $Y$ are blocked and any residual
association between $T$ and $Y$ conditional on $\mathbf{S}$ is purely
causal.  By the truncated factorization formula for $P(\cdot\mid\doop(t))$,
the post-intervention conditional distribution of $Y$ given $T=t$ and
$\mathbf{S}=\mathbf{s}$ therefore equals the observational conditional:
\[
  P(y \mid \doop(t), \mathbf{s}) = P(y \mid t, \mathbf{s}).
\]
(\textit{Note}: this step does \emph{not} follow from Rule~2 directly,
because the Rule~2 condition $(Y \indep T \mid \mathbf{S})_{\Gcal_{\overline{T(\mathbf{S})}}}$
fails whenever $T \to Y$ is a direct edge.  The equality instead holds by
the structural argument above: condition~2 implies
$(Y \indep T \mid \mathbf{S})_{\Gcal_{\underline{T}}}$, which together
with the truncated factorization gives the result.)

\noindent\textbf{Combining:}
\[
  P(y \mid \doop(t))
  = \int P(y \mid t, \mathbf{s})\,dP(\mathbf{s}).
  \hfill\square
\]
\end{tcolorbox}

\begin{tcolorbox}[defstyle, title={Proof of Theorem~\ref{w3:thm:frontdoor} (Front-Door Formula) via Three Applications of the Rules}]

\noindent\textit{Graph structure}: $T \to M \to Y$, unobserved $U \to T$,
$U \to Y$, no direct $T \to Y$ edge.

\begin{center}
\begin{tikzpicture}[node distance=1.6cm]
  \node[node]  (T) at (0, 0)     {$T$};
  \node[node]  (M) at (2.2, 0)   {$M$};
  \node[node]  (Y) at (4.4, 0)   {$Y$};
  \node[unode] (U) at (2.2, 1.4) {$U$};
  \draw[edge]  (T) -- (M) node[midway, below,
    font=\scriptsize\itshape\color{darkgrey}] {Stage 1};
  \draw[edge]  (M) -- (Y) node[midway, below,
    font=\scriptsize\itshape\color{darkgrey}] {Stage 2};
  \draw[dedge] (U) -- (T);
  \draw[dedge] (U) -- (Y);
\end{tikzpicture}
\end{center}

\noindent Stage~1 ($T \to M$) is unconfounded by condition~2.
Stage~2 ($M \to Y$) is confounded by $U$ via $M \leftarrow T \leftarrow U
\to Y$, but conditioning on $T$ blocks this path (condition~3).

\medskip
\noindent\textbf{Step 1: Introduce $M$ by the law of total probability.}
\[
  P(y \mid \doop(t)) = \int P(y \mid \doop(t),\;m)\,dP(m \mid \doop(t)).
\]

\noindent\textbf{Step 2: Simplify $P(m \mid \doop(t))$.}
In $\Gcal_{\overline{T}}$, there are no back-door paths from $T$ to $M$
by front-door condition~2.  Rule~2 applies:
\[
  P(m \mid \doop(t)) = P(m \mid t).
\]

\noindent\textbf{Step 3: Simplify $P(y \mid \doop(t), m)$.}

\medskip
\noindent\textit{Step 3a}: $P(y \mid \doop(t), m) = P(y \mid \doop(m))$.

In $\Gcal_{\overline{T}}$, the arrow $U \to T$ is deleted, so $U$ and $T$
become independent.  Since there is no arrow from $U$ into $M$ (condition~2),
$U$ is also independent of $M$ in $\Gcal_{\overline{T}}$.  Therefore:
\[
  P(y \mid \doop(t), m) \;=\; \int P(y \mid m, u)\,dP(u) \;=\; P(y \mid \doop(m)).
\]

\medskip
\noindent\textit{Step 3b}: $P(y \mid \doop(m)) = \int P(y \mid m, t')\,dP(t')$.

By front-door condition~3, $\{T\}$ satisfies the back-door criterion for
the effect of $M$ on $Y$, and Theorem~\ref{w3:thm:backdoor} gives:
\[
  P(y \mid \doop(m)) \;=\; \int P(y \mid m, t')\,dP(t').
\]

\noindent Combining Steps 3a and 3b:
\[
  P(y \mid \doop(t), m) \;=\; \int P(y \mid m, t')\,dP(t').
\]

\noindent\textbf{Assembling:}
\[
  P(y \mid \doop(t))
  = \int\!\left[\int P(y \mid t',\,m)\,dP(t')\right]dP(m \mid t).
  \hfill\square
\]
\end{tcolorbox}

% ==============================================================
\section{The Do-Calculus is Complete}
\label{w3:sec:completeness}
% ==============================================================

\noindent\textbf{General theory (continued).}\enspace
This section is included to explain the scope and limits of the
do-calculus.  Students need not master the ID algorithm procedurally on
a first reading; the main takeaway is that identifiability is a structural
property of the graph, not merely a matter of algebraic cleverness.

A natural question after seeing the three rules is: are they enough?
Could there be a graph where the causal effect is identifiable in principle,
but none of the rules --- applied in any sequence --- can derive a purely
observational expression?  \citet{shpitser2006identification} answered this
definitively.

Throughout Sections~\ref{w3:sec:mutilated}--\ref{w3:sec:rules} we have
represented latent common causes as
explicit unobserved nodes (e.g.\ $U \to T$ and $U \to Y$).  The
completeness theorem uses a more general graph class that encodes latent
common causes compactly as \emph{bidirected edges}: a bidirected edge
$X \leftrightarrow Y$ stands for an unobserved variable $U$ with $U \to X$
and $U \to Y$.  A DAG augmented with such bidirected edges is called a
\emph{semi-Markovian DAG}.  The IV DAG and front-door graphs of this course
are both special cases.

\begin{thm}{Completeness of the Do-Calculus \citep{shpitser2006identification}}
\label{w3:thm:completeness}
Let $\Gcal$ be a semi-Markovian DAG.  A causal quantity
$P(y \mid \doop(t))$ is identifiable from $P$ relative to $\Gcal$ if and
only if the do-calculus can derive a purely observational expression for it.
\end{thm}

The proof is beyond the scope of this chapter; the sketch below is
provided for reference and independent reading.  Before turning to the
proof machinery, we ground the theorem in the simplest possible
non-identifiable graph.

\begin{example}{The Bow Graph: The Simplest Non-Identifiable Structure}
\label{w3:ex:bowgraph}
The \emph{bow graph} is the semi-Markovian DAG with one directed edge
$T \to Y$ and one bidirected edge $T \leftrightarrow Y$ (representing an
unobserved $U$ with $U \to T$ and $U \to Y$):

\begin{center}
\begin{tikzpicture}[node distance=2.0cm]
  %--- left panel: explicit U node ---
  \node[node]  (T1) at (0, 0)     {$T$};
  \node[node]  (Y1) at (2.4, 0)   {$Y$};
  \node[unode] (U1) at (1.2, 1.2) {$U$};
  \draw[edge]  (T1) -- (Y1);
  \draw[dedge] (U1) -- (T1);
  \draw[dedge] (U1) -- (Y1);
  \node[font=\small\itshape\color{darkgrey}] at (1.2,-0.7)
    {explicit $U$};

  %--- right panel: bidirected notation ---
  \node[node]  (T2) at (5.2, 0)   {$T$};
  \node[node]  (Y2) at (7.6, 0)   {$Y$};
  \draw[edge]  (T2) -- (Y2);
  \draw[{Stealth[length=5pt]}-{Stealth[length=5pt]},
        thick, color=darkgrey, dashed,
        bend left=45] (T2) to (Y2);
  \node[font=\small\itshape\color{darkgrey}] at (6.4,-0.7)
    {semi-Markovian notation};
\end{tikzpicture}
\end{center}

\noindent
The single back-door path is $T \leftarrow U \to Y$.  Because $U$ is
unobserved and there are no other variables in the graph, no observed set
can block this path, so the back-door criterion cannot be satisfied.  The
front-door criterion also fails: there is no observed mediator on any
directed path from $T$ to $Y$.  We now show \emph{directly} that
$P(y \mid \doop(t))$ is not determined by the observed distribution
$P(T, Y)$ alone, by constructing two models that agree on $P(T, Y)$ but
disagree on $P(y \mid \doop(t))$.

Let $T, Y, U \in \{0, 1\}$ with $U \sim \mathrm{Bernoulli}(1/2)$.

\medskip
\noindent\textbf{Model $\mathcal{M}_1$ (pure confounding; no causal effect of $T$).}
\[
  T \;=\; U, \qquad Y \;=\; U.
\]
Both $T$ and $Y$ are driven entirely by $U$; the arrow $T \to Y$ carries
no causal influence.  Since $T = U$ and $Y = U$, the observed joint
distribution is $P(T{=}0, Y{=}0) = P(T{=}1, Y{=}1) = 1/2$ and
$P(T \neq Y) = 0$.

Under the intervention $\doop(T{=}1)$, the equation $T = U$ is replaced
by the constant $T = 1$, but $Y = U$ is unchanged.  Therefore:
\[
  P_1\!\left(Y{=}1 \mid \doop(T{=}1)\right) \;=\; P(U{=}1) \;=\; \tfrac{1}{2}.
\]

\medskip
\noindent\textbf{Model $\mathcal{M}_2$ (full causal effect; $U$ acts on $Y$ only through $T$).}
\[
  T \;=\; U, \qquad Y \;=\; T.
\]
$Y$ is determined entirely by $T$; $U$ affects $Y$ only via $T$.  Since
$T = U$ again, we have $Y = T = U$, so the observed joint distribution is
again $P(T{=}0, Y{=}0) = P(T{=}1, Y{=}1) = 1/2$ and $P(T \neq Y) = 0$
--- \emph{identical} to $\mathcal{M}_1$.

Under the intervention $\doop(T{=}1)$, the equation $T = U$ is replaced
by $T = 1$, and $Y = T = 1$ deterministically.  Therefore:
\[
  P_2\!\left(Y{=}1 \mid \doop(T{=}1)\right) \;=\; 1.
\]

\medskip
\noindent\textbf{Conclusion.}  The two models produce identical observed
distributions $P(T,Y)$, so no observational data set, however large, can
distinguish $\mathcal{M}_1$ from $\mathcal{M}_2$.  Yet they assign
different values to $P(Y{=}1 \mid \doop(T{=}1))$ --- $1/2$ in
$\mathcal{M}_1$ versus $1$ in $\mathcal{M}_2$ --- and correspondingly
different average causal effects ($0$ versus $1$).  Therefore
$P(y \mid \doop(t))$ is \emph{not identified} from $P(T, Y)$ in the bow
graph.  The completeness theorem tells us that no identification method
--- the do-calculus or otherwise --- can overcome this obstruction: the
bow graph contains a hedge, and the ID algorithm returns
\textsc{fail}.
\end{example}

\begin{tcolorbox}[defstyle, title={Proof Sketch and the ID Algorithm}]
\citet{shpitser2006identification} introduced the \emph{ID algorithm}
(Identification algorithm): a procedure that takes as input a
semi-Markovian DAG $\Gcal$, a target intervention distribution
$P(y \mid \doop(t))$, and the observed joint distribution $P$, and either
returns an explicit observational formula for $P(y \mid \doop(t))$ or
reports that the quantity is not identifiable from $P$ given $\Gcal$.

\textbf{Sufficiency} (do-calculus identifies whenever identification is
possible): proved constructively via the ID algorithm.  The key structural
concept is the \emph{c-component} (confounded component): a maximal set of
nodes connected by bidirected edges, representing a group of variables that
share latent common causes.  For a single node $V$ with no bidirected
edges, its c-component is $\{V\}$ itself.

The ID algorithm proceeds recursively as follows.
\begin{enumerate}
  \item \textit{Simplification}: remove from $\Gcal$ any node that is not
        an ancestor of $Y$ in $\Gcal_{\overline{T}}$.  Such nodes cannot
        affect $P(y \mid \doop(t))$ and can be marginalized out.
  \item \textit{Decomposition by c-components}: factorise
        $P(\mathbf{v} \setminus \mathbf{t} \mid \doop(t))$ over the
        c-components of $\Gcal_{\overline{T}}$.  Each factor depends only
        on the variables in one c-component and their ancestors, reducing
        the problem to smaller subproblems.
  \item \textit{Recursive identification}: attempt to identify each
        c-component factor from $P$ using the do-calculus rules and the
        Markov factorization.  If every factor is successfully reduced to
        an observational expression, assemble the results to obtain the
        final formula for $P(y \mid \doop(t))$.
  \item \textit{Failure}: if any subproblem cannot be reduced --- because
        the c-component of $Y$ is not identifiable given the surrounding
        structure --- the algorithm returns \textsc{fail}.
\end{enumerate}
When the algorithm succeeds, it has produced an explicit identification
formula, demonstrating that the do-calculus is sufficient.

\textbf{Necessity} (do-calculus cannot identify non-identifiable
quantities): proved by showing that whenever the ID algorithm fails, the
causal effect is genuinely non-identifiable --- not merely hard to compute.
The key graphical obstruction is the \emph{hedge}.

\begin{quote}
\textit{A \emph{hedge} for $P(y \mid \doop(t))$ in $\Gcal$ is a pair of
subgraphs $(F, F')$ where $F$ and $F'$ are both rooted at the same set of
nodes, $F$ is a subgraph of $\Gcal$ containing a back-door path from $T$
to $Y$, and $F'$ is a subgraph of $\Gcal_{\overline{T}}$ in which $Y$
remains connected to a latent common cause.  Intuitively, the hedge
encodes a ``loop'' of confounding that the do-calculus cannot break.}
\end{quote}

\citet{shpitser2006identification} prove that the ID algorithm fails if and
only if $\Gcal$ contains a hedge for $P(y \mid \doop(t))$.  When a hedge
exists, one can explicitly construct two semi-Markovian models
$\mathcal{M}_1$ and $\mathcal{M}_2$ that:
\begin{itemize}
  \item induce the same observed joint distribution $P$, so no data can
        distinguish them; but
  \item assign different values to $P(y \mid \doop(t))$, so the
        interventional distribution is not a function of $P$ alone.
\end{itemize}
The construction builds $\mathcal{M}_1$ and $\mathcal{M}_2$ by choosing
different structural equations for the latent variables along the hedge,
keeping all observed conditional distributions identical.  Since the two
models are observationally indistinguishable yet causally distinct, no
method --- including the do-calculus --- can identify $P(y \mid \doop(t))$
from $P$ alone.  This establishes necessity.

Combining sufficiency and necessity: the do-calculus succeeds if and only
if no hedge exists, which is if and only if $P(y \mid \doop(t))$ is
identifiable. $\square$
\end{tcolorbox}

\begin{remark}
Theorem~\ref{w3:thm:completeness} has a crucial practical implication:
\emph{if the ID algorithm reports non-identification, then no estimation
method --- however clever --- can recover the causal effect from
observational data alone}, given the assumed graph structure.
Non-identification is not a limitation of a particular technique; it is a
fundamental property of the causal model.  Three remedies exist: impose
additional structural assumptions (e.g., linearity or monotonicity) that
go beyond the graph; collect additional data that break the obstruction
(e.g., an instrument or a randomized experiment); or target a different,
identified causal quantity instead (e.g., partial identification bounds,
or the effect among the treated rather than the average treatment effect).
\end{remark}

% ==============================================================
\section{The Big Picture}
\label{w3:sec:bigpicture}
% ==============================================================

This chapter completes the \emph{theoretical} identification machinery of
the course.  The chain of reasoning begun in Chapter~1 (the
interventional/observational distinction) and formalised in Chapter~2
(d-separation) now terminates in a concrete identification formula.
Chapters~4--7 then apply this machinery to specific research designs:

\bigskip
\begin{center}
\begin{tikzpicture}[
    box/.style = {
      rectangle, rounded corners=6pt,
      draw=isublue, fill=defbg, thick,
      text width=7.2cm, align=center,
      minimum height=1.1cm,
      font=\bfseries\color{isublue}
    },
    tool/.style = {
      rectangle, rounded corners=3pt,
      draw=accent, fill=white, thin,
      text width=4.8cm, align=center,
      font=\small\itshape\color{accent}
    },
    arr/.style = {-{Stealth[length=7pt,width=5pt]},
                  line width=1.6pt, color=accent},
    node distance = 0.55cm
  ]
  \node[box] (G)  {Graph Structure ($\Gcal$)};
  \node[box, below=1.2cm of G]  (D)
    {Mutilated Graphs ($\Gcal_{\overline{T}},\, \Gcal_{\underline{T}}$)};
  \node[box, below=1.2cm of D]  (R)
    {Do-Calculus Rules (1, 2, 3)};
  \node[box, below=1.2cm of R]  (ID)
    {Identification Formula $P(y\mid\doop(t)) = \Psi(P_{\mathrm{obs}})$};
  \node[box, below=1.2cm of ID] (W46)
    {Research Designs: Potential Outcomes, Mediation, Propensity Scores, IV
     (Chapters~4--8)};
  \node[box, below=1.2cm of W46] (ES)
    {Estimation: $\hat\Psi_n$ (Chapters~9--12)};

  \draw[arr] (G)   -- (D);
  \draw[arr] (D)   -- (R);
  \draw[arr] (R)   -- (ID);
  \draw[arr] (ID)  -- (W46);
  \draw[arr] (W46) -- (ES);

  \node[tool, right=1.1cm of G]
    {Causal assumptions encoded as directed edges};
  \node[tool, right=1.1cm of D]
    {$\Gcal_{\overline{T}}$ models intervention;
     $\Gcal_{\underline{T}}$ is a technical device for Rule~1};
  \node[tool, right=1.1cm of R]
    {Three rules license algebraic transformations};
  \node[tool, right=1.1cm of ID]
    {Back-door, front-door, or ID algorithm};
  \node[tool, right=1.1cm of W46]
    {Apply identification framework to concrete designs};
  \node[tool, right=1.1cm of ES]
    {GMM / EIF / doubly robust (Part~III)};

  \foreach \n in {G, D, R, ID, W46, ES}
    \draw[dotted, color=accent!50, thin] (\n.east) -- ++(1.1cm,0);
\end{tikzpicture}
\end{center}

\medskip
Chapter~4 introduces the potential outcomes framework and its connection
to the do-calculus, providing the second language for causal inference.
Chapters~5--6 apply these tools to randomization and back-door adjustment,
and to propensity score methods.  Chapter~7 develops instrumental variables
as the primary strategy when unmeasured confounding rules out the back-door
route.  Chapter~8 then studies mediation analysis, decomposing total effects
into direct and indirect components.
The completeness theorem (Section~\ref{w3:sec:completeness}) closes the
loop: every causal effect that is identifiable from the observational
distribution relative to the assumed graph can be derived by the
do-calculus, and every effect that is not identifiable in that setting
cannot be recovered by any observational method without additional
assumptions or new data.  Part~III (Chapter~9 and beyond) then
turns to estimation.

\medskip

% ==============================================================
\section{Summary}
\label{w3:sec:summary}
% ==============================================================

\begin{enumerate}
  \item Two graph operations underpin the do-calculus:
        $\Gcal_{\overline{X}}$ (delete arrows \emph{into} $X$, modeling
        the intervention $\doop(x)$) and $\Gcal_{\underline{X}}$ (delete
        arrows \emph{out of} $X$, used as a technical device to check the
        d-separation condition for Rule~1 --- it does not model
        conditioning on $X$).

  \item The back-door criterion identifies $P(y \mid \doop(t))$ when an
        observed set $\mathbf{S}$ blocks all back-door paths and contains
        no descendant of $T$.  The general formula is
        $\int P(y \mid t, \mathbf{s})\,dP(\mathbf{s})$,
        equivalently $\E_{\mathbf{S}}\!\left[\E[h(Y)\mid T{=}t,\mathbf{S}]\right]$
        in iterated-expectation form (standardization; the g-formula for
        $K=1$).

  \item The front-door criterion identifies $P(y \mid \doop(t))$ via an
        observed mediator $M$ intercepting all causal paths from $T$ to
        $Y$, even when the confounder $U$ is entirely unobserved, provided
        the three front-door conditions hold.

  \item The do-calculus has three rules, each licensed by a d-separation
        condition in an appropriate intervention graph.  Rule~1 adds/removes
        observations; Rule~2 exchanges actions for observations; Rule~3
        deletes actions.  The back-door and front-door formulas are each
        derivable as three-step sequences of these rules
        (Section~\ref{w3:sec:proofs}).

  \item \textbf{Completeness.} The do-calculus is complete for
        identification from observational data relative to the assumed
        graph \citep{shpitser2006identification}: a causal effect is
        identifiable if and only if the do-calculus can derive a purely
        observational expression for it.  If the do-calculus cannot
        identify the quantity in that setting, then no purely
        observational method can do so without additional assumptions or
        new data.
\end{enumerate}

% ==============================================================
\newpage
\section*{Problems}
\addcontentsline{toc}{section}{Problems}
% ==============================================================

\begin{enumerate}[label=\textbf{\arabic*.}, leftmargin=*, itemsep=10pt]

  \item \textbf{Warm-up: graph operations.}
  Let $\Gcal$ be the DAG with edges $Z \to T$, $T \to Y$, $U \to T$,
  $U \to Y$ (the IV DAG, with $U$ unobserved).
  \begin{enumerate}[label=(\alph*)]
    \item Draw $\Gcal_{\overline{T}}$ and identify all paths from $Z$ to
          $Y$ that remain open.
    \item Draw $\Gcal_{\underline{T}}$ and identify all paths from $Z$ to
          $Y$ that remain.  For each path, classify it as a causal chain,
          fork, or collider path, and state whether it is open or closed
          by default (without conditioning on any variable).
    \item In $\Gcal_{\overline{T}}$, is $Z \indep Y \mid \varnothing$?
          How does the absence of a direct edge $Z \to Y$ in the DAG
          relate to the exclusion restriction (the assumption that $Z$
          affects $Y$ only through $T$)?
  \end{enumerate}

  \item \textbf{Back-door practice.}
  Consider the DAG: $X \to T$, $X \to Y$, $T \to M$, $M \to Y$,
  $T \to Y$, where $X$ is observed.
  \begin{enumerate}[label=(\alph*)]
    \item List all back-door paths from $T$ to $Y$.
    \item Does $\{X\}$ satisfy the back-door criterion for the effect of
          $T$ on $Y$?  Write the resulting adjustment formula.
    \item Does $\{M\}$ satisfy the back-door criterion?  Explain why or
          why not.
    \item Does $\{X, M\}$ satisfy the back-door criterion?  Identify
          which condition of the criterion $M$ violates, and explain why
          this rules it out as an adjustment variable.
    \item Even if the formal criterion issue in (d) were somehow set aside,
          explain the substantive bias that arises from conditioning on a
          mediator $M$ that lies on a causal path $T \to M \to Y$.
  \end{enumerate}

  \item \textbf{Front-door identification.}
  Consider the DAG: $T \to M \to Y$, with $U \to T$ and $U \to Y$
  (unobserved $U$, no direct $T \to Y$ edge).
  \begin{enumerate}[label=(\alph*)]
    \item Verify that $M$ satisfies all three front-door conditions.
    \item Derive the front-door formula~\eqref{w3:eq:frontdoor} step by
          step, citing the do-calculus rule used at each step.
    \item Now add a direct edge $T \to Y$ to the graph.  Does $M$ still
          satisfy the front-door criterion?  Explain which condition fails.
  \end{enumerate}

  \item \textbf{Identification or non-identification?}
  For each of the following DAGs, determine whether
  $P(y \mid \doop(t))$ is identified \emph{non-parametrically} from
  observational data.  If identified, state the formula; if not, explain
  the structural obstruction.
  \begin{enumerate}[label=(\alph*)]
    \item $T \to Y$, $U \to T$, $U \to Y$, $U$ unobserved.
    \item $T \to Y$, $U \to T$, $U \to Y$, $U$ unobserved; with
          instrument $Z \to T$ and $Z \indep U$ (no direct $Z \to Y$ edge).
          Show that $P(y \mid \doop(t))$ is \textbf{not} identified
          non-parametrically from $P(Z, T, Y)$ by constructing two binary
          SEMs $\mathcal{M}_1$ and $\mathcal{M}_2$, each compatible with
          the IV graph, such that (i)~$P_{\mathcal{M}_1}(Z, T, Y) =
          P_{\mathcal{M}_2}(Z, T, Y)$ and (ii)~$P_{\mathcal{M}_1}(Y{=}1
          \mid \doop(T{=}1)) \ne P_{\mathcal{M}_2}(Y{=}1 \mid
          \doop(T{=}1))$.  (\textit{Hint}: let $Z, U \sim
          \mathrm{Bern}(1/2)$ independently and set $T = U$ in both
          models --- the IV graph permits $Z$ to enter $T$'s structural
          equation with zero coefficient.  Then set $Y = U$ in
          $\mathcal{M}_1$ and $Y = T$ in $\mathcal{M}_2$.  Verify that
          both models produce the same joint $P(Z, T, Y)$, and compute
          the ACE under $\doop(T{=}1)$ in each model.})
    \item $T \to M \to Y$, $U \to T$, $U \to Y$, $U \to M$, $U$
          unobserved.  (\textit{Hint}: first check whether the back-door
          criterion can be satisfied --- is there any observed variable
          that blocks the path $T \leftarrow U \to Y$?  Then check the
          front-door criterion: does $M$ satisfy condition~2?  Conclude
          whether the effect is identified by either strategy.)
    \item $T \to Y$, $X \to T$, $X \to Y$, $X$ observed.
  \end{enumerate}

  \item \textbf{Back-door formula: proof and uniqueness.}

  \begin{enumerate}[label=(\alph*)]
    \item Let $\mathcal{M}$ be a structural causal model Markov with
          respect to $\Gcal$, and let $\mathbf{S}$ satisfy the back-door
          criterion for $(T, Y)$ in $\Gcal$ with positivity.  Prove
          that
          \[
            P(y \mid \doop(t))
            \;=\; \int P(y \mid t,\, \mathbf{s})\,dP(\mathbf{s}).
          \]
          (\textit{Hint}: begin with
          $P(y \mid \doop(t)) = \int P(y \mid \doop(t), \mathbf{s})
          \,P(\mathbf{s} \mid \doop(t))\,d\mathbf{s}$.
          Apply the Rule~3 graphical condition
          $(\mathbf{S} \indep T)_{\Gcal_{\overline{T}}}$, using
          back-door condition~1, to simplify $P(\mathbf{s} \mid
          \doop(t))$.  Then invoke the d-separation
          $(Y \indep T \mid \mathbf{S})_{\Gcal_{\underline{T}}}$,
          which back-door condition~2 implies, together with the
          truncated factorization, to argue
          $P(y \mid \doop(t), \mathbf{s}) = P(y \mid t,
          \mathbf{s})$.)

    \item Suppose $\mathbf{S}_1$ and $\mathbf{S}_2$ both satisfy the
          back-door criterion for $(T, Y)$ in $\Gcal$ (with
          positivity).  Deduce from part~(a) that
          \[
            \int P(y \mid t,\, \mathbf{s}_1)\,dP(\mathbf{s}_1)
            \;=\;
            \int P(y \mid t,\, \mathbf{s}_2)\,dP(\mathbf{s}_2).
          \]
          Interpret this result: the identification target
          $P(y \mid \doop(t))$ is unique even when the adjustment set
          is not.
  \end{enumerate}

  \item \textbf{Rule sequencing on an unseen graph.}
  Consider the graph from Example~\ref{w3:ex:twopaths}:
  $C_1 \to T$, $C_1 \to Y$, $U \to T$, $U \to C_2$, $C_2 \to Y$,
  $T \to Y$, with $C_1$ and $C_2$ observed and $U$ unobserved.
  Example~\ref{w3:ex:twopaths} established that $\{C_1, C_2\}$ is a
  valid back-door adjustment set.

  Derive
  $P(y \mid \doop(t)) = \sum_{c_1,c_2}
  P(y \mid t, c_1, c_2)\,P(c_1, c_2)$
  from first principles using the three rules.  For each step:
  (i)~name the rule applied,
  (ii)~form the relevant intervention graph by stating which arrows are
       deleted and what edges remain, and
  (iii)~verify the required d-separation condition by tracing all paths
        and confirming each is blocked.

  \item \textbf{Non-identifiability by construction.}
  This problem practices the technique of
  Example~\ref{w3:ex:bowgraph}: prove non-identifiability by
  exhibiting two observationally equivalent models with different
  causal effects.

  Consider the graph from Case~(a) of
  Example~\ref{w3:ex:frontdoorfail}: $T \to M \to Y$,
  $U \to T$, $U \to M$, $U \to Y$ (front-door condition~2 violated
  because $U$ confounds $T \to M$).  Let $T, M, Y, U \in \{0,1\}$
  with $U \sim \mathrm{Bern}(1/2)$.

  \begin{enumerate}[label=(\alph*)]
    \item Construct two SEMs $\mathcal{M}_1$ and $\mathcal{M}_2$,
          each compatible with the graph, such that
          $P_{\mathcal{M}_1}(T, M, Y) = P_{\mathcal{M}_2}(T, M, Y)$
          but
          $P_{\mathcal{M}_1}(Y{=}1 \mid \doop(T{=}1))
          \ne P_{\mathcal{M}_2}(Y{=}1 \mid \doop(T{=}1))$.
          (\textit{Hint}: set $T = U$ in both models.  In
          $\mathcal{M}_1$ set $M = U$ and $Y = U$; in $\mathcal{M}_2$
          set $M = T$ and $Y = M$.  Verify both produce
          $P(T{=}0, M{=}0, Y{=}0) = P(T{=}1, M{=}1, Y{=}1) =
          \tfrac{1}{2}$, then compute the ACE in each model under
          $\doop(T{=}1)$.)

    \item Explain precisely which identification strategy fails in
          this graph and why.  Your explanation should reference
          (i)~the back-door criterion (is there any observed variable
          that blocks $T \leftarrow U \to Y$?), (ii)~the front-door
          criterion (which condition fails and which stage of the
          two-stage argument collapses), and (iii)~the completeness
          theorem (what does the existence of your two models imply
          about any identification method?).
  \end{enumerate}

\end{enumerate}

% ==============================================================
\newpage
\ifSubfilesClassLoaded{
  \bibliographystyle{plainnat}
  \bibliography{causal_inference}
}{}

\end{document}