\documentclass[master.tex]{subfiles}

\begin{document}

% ======================== TITLE ================================
\ifSubfilesClassLoaded{%
  % Standalone title block
  \begin{center}
    {\LARGE\bfseries\color{isublue} Chapter 2: DAGs and d-Separation}\\[6pt]
    {\large Theory and Methods in Causal Inference}\\[4pt]
    {\normalsize Department of Statistics, Iowa State University}
  \end{center}
  \bigskip
  \hrule height 1.2pt \relax
  \smallskip
}{%
  % Master book: chapter heading
  \chapter{DAGs and d-Separation}
}
% ======================== LEARNING OBJECTIVES ==================
\begin{tcolorbox}[defstyle, title={Learning Objectives}]
By the end of this chapter, students should be able to:
\begin{enumerate}
  \item Construct a DAG from a verbal description of a causal model and identify
        parents, descendants, ancestors, and non-descendants of any node.
  \item Classify every intermediate node on a path as a chain, fork, or collider,
        and apply the blocking rule for each.
  \item Apply the d-separation criterion to determine all conditional independence
        relationships implied by a DAG.
  \item Recognize and avoid collider bias, including the case where a descendant
        of a collider is conditioned upon.
  \item State the Markov factorization and use it to write the joint density of a
        DAG in terms of local conditional densities.
  \item Interpret d-separation results as conditional independence assumptions
        on observed variables, and recognize their role as the graphical
        expression of causal assumptions.
\end{enumerate}
\end{tcolorbox}

% ---------------------------------------------------------------
% Styles and macros specific to this chapter (not yet in master).
% Placed here (after the LO box, before §1) so the definition block
% does not inject blank-line spacing between \chapter{} and the box.
% The only \begin{remark} in this chapter is in §4, well below.
% (remarkstyle and remark environment are defined in causal_inference_master.tex)

% ==============================================================
\section{Directed Acyclic Graphs}
% ==============================================================

DAGs allow us to translate qualitative causal assumptions into quantitative
statistical restrictions.  This single idea underlies everything in this
chapter: once we draw a graph encoding what we believe about the causal
structure of a problem, the graph tells us exactly which variables are
independent, which are confounded, and what we need to condition on to
identify a causal effect.

\begin{remark}
In this chapter, the word \emph{graph} does not mean a plot of data or
a graph of a function.
It means a collection of nodes and edges used to represent relationships
among variables.
Our objective is to understand how such graphs encode statistical
structure --- especially conditional independence, confounding, and path
blocking --- and how this graphical structure will later support causal
identification.
\end{remark}

\subsection{Basic Definition}

\begin{defn}{Directed Acyclic Graph}
A \emph{directed acyclic graph} (DAG) $\Gcal = (V, E)$ consists of a finite
set of nodes $V$ and a set of directed edges $E \subseteq V \times V$ such that there
is no directed cycle: no node $V_i$ has a directed path to itself.
\end{defn}

In a causal DAG each node represents a variable and each directed edge $A \to B$
encodes ``$A$ is a direct cause of $B$, given all other variables in the model.''
The acyclicity condition is essential for the standard DAG semantics.
In particular, it guarantees that the variables can be ordered topologically
so that the joint density admits the recursive factorization
\[
  p(v_1,\dots,v_k) = \prod_{i=1}^{k} p(v_i \mid \Pa(v_i)).
\]
If the graph contained a directed cycle, this DAG factorization would no longer
be available.  Cyclic causal systems may still be studied in other frameworks,
but they are not DAGs and are outside the scope of these notes.

\subsection{Structural Relationships}

\begin{defn}{Structural relationships in a DAG}
For a node $v \in \mathcal{V}$ in $\Gcal = (\mathcal{V}, E)$:
\begin{itemize}
  \item $\Pa(v) = \{w \in \mathcal{V} : w \to v \in E\}$ are the \emph{parents} of $v$.
  \item $\De(v)$ = all nodes reachable from $v$ by directed paths
        are the \emph{descendants} of $v$.
  \item $\An(v)$ = all nodes with a directed path to $v$ are the
        \emph{ancestors} of $v$.
  \item $\Nd(v) = \mathcal{V} \setminus (\De(v) \cup \{v\})$ are the
        \emph{non-descendants} of $v$.
\end{itemize}
Note that $\Nd(v)$ includes the parents of $v$; when stating the local
Markov property in Section~\ref{w2:sec:markov}, we will exclude the parents
explicitly.
\end{defn}

\begin{tcolorbox}[defstyle, title={Quick Terminology Summary}]
A \emph{node} represents a variable, and an \emph{edge} represents a
relationship between two variables.
If $A \to B$, then $A$ is a \emph{parent} of $B$ and $B$ is a
\emph{child} of $A$.
Two nodes are \emph{adjacent} if they are connected by an edge.
A \emph{path} is any sequence of connected nodes, regardless of arrow
direction; a \emph{directed path} is a path whose arrows all point in
the same direction.
A node is an \emph{ancestor} of another if there is a directed path from
the former to the latter, and a \emph{descendant} if the direction is
reversed.
A \emph{cycle} is a directed path that returns to its starting node.
A \emph{directed acyclic graph} (DAG) is a directed graph with no
directed cycles.
\end{tcolorbox}

\begin{example}{Education, Earnings, and Family Background}
Consider the causal story: education ($E$) affects earnings ($Y$); family
background ($B$) is a common cause of both; and neighborhood ($N$) affects
education but has no direct effect on earnings.

\begin{center}
\begin{tikzpicture}[node distance=1.8cm]
  \node[node] (N) {$N$};
  \node[node, right of=N] (E) {$E$};
  \node[node, right of=E] (Y) {$Y$};
  \node[node, below=1.0cm of E] (B) {$B$};
  \draw[edge] (N) -- (E);
  \draw[edge] (E) -- (Y);
  \draw[edge] (B) -- (E);
  \draw[edge] (B) -- (Y);
\end{tikzpicture}
\end{center}

The Markov factorization is $p(n,b,e,y) = p(n)\,p(b)\,p(e\mid n,b)\,p(y\mid e,b)$.
Reading off structural relationships: $\Pa(E) = \{N,B\}$,
$\Pa(Y) = \{E,B\}$, $\Pa(N)=\Pa(B)=\varnothing$.
The descendants of $N$ are $\{E,Y\}$; $B$ and $N$ are non-descendants of each other.
\end{example}

\begin{example}{The IV DAG}
The canonical instrumental variable model involves four nodes: $Z$ (instrument),
$T$ (treatment), $Y$ (outcome), and $U$ (unobserved confounder).

\begin{center}
\begin{tikzpicture}[node distance=1.8cm]
  \node[node] (Z) {$Z$};
  \node[node, right of=Z] (T) {$T$};
  \node[node, right of=T] (Y) {$Y$};
  \node[unode, below=1.0cm of T] (U) {$U$};
  \draw[edge] (Z) -- (T);
  \draw[edge] (T) -- (Y);
  \draw[dedge] (U) -- (T);
  \draw[dedge] (U) -- (Y);
\end{tikzpicture}
\end{center}

$\Pa(T) = \{Z,U\}$, $\Pa(Y) = \{T,U\}$,
$\Pa(Z) = \Pa(U) = \varnothing$.
The descendants of $Z$ are $\{T, Y\}$; $U$ and $Z$ are non-descendants of each other.
\end{example}

\begin{example}{The Mediation DAG}
In a mediation model, treatment $T$ affects outcome $Y$ both directly and
through an intermediate variable $M$ (the mediator).  An unobserved confounder
$U$ creates a back-door path $T \leftarrow U \rightarrow Y$, making $T$
endogenous.  The mediator $M$ is itself unconfounded.

\begin{center}
\begin{tikzpicture}[node distance=1.8cm]
  \node[node] (T) {$T$};
  \node[node, right of=T] (M) {$M$};
  \node[node, right of=M] (Y) {$Y$};
  \node[unode, below=1.0cm of T] (U) {$U$};
  \draw[edge] (T) -- (M);
  \draw[edge] (M) -- (Y);
  \draw[edge] (T) to[bend left=30] (Y);
  \draw[dedge] (U) -- (T);
  \draw[dedge] (U) -- (Y);
\end{tikzpicture}
\end{center}

$\Pa(T) = \{U\}$, $\Pa(M) = \{T\}$,
$\Pa(Y) = \{M, T, U\}$, $\Pa(U) = \varnothing$.
The total effect of $T$ on $Y$ travels along two paths: the direct path
$T \to Y$ and the indirect path $T \to M \to Y$.
This graph is a standard mediation DAG, not a front-door DAG.
Because there is a direct path $T \to Y$ bypassing $M$, the front-door
criterion does not apply here.
A front-door graph would require that every directed path from $T$ to $Y$
pass through $M$.
\end{example}

% ==============================================================
\section{Paths, Blocking, and d-Separation}
% ==============================================================

A DAG encodes direct causal relationships through its edges, but variables
can also be statistically related through longer routes that mix directed
and reversed edges.  The concept of a \emph{path} formalises these routes,
and the question of whether a path transmits or blocks statistical
dependence --- depending on which variables are conditioned upon --- is
precisely what d-separation answers.  The key insight is that conditioning
can both \emph{close} channels that were open (by blocking a chain or fork)
and \emph{open} channels that were closed (by activating a collider).
Figure~\ref{w2:fig:chain-fork-collider} illustrates all three cases.

\subsection{Path Structures}

\begin{defn}{Path}
A \emph{path} between nodes $X$ and $Y$ in $\Gcal$ is any sequence of
distinct nodes $X = V_0, V_1, \dots, V_k = Y$ such that each consecutive pair
is connected by an edge in either direction.
\end{defn}

Every intermediate node $V_m$ on a path plays one of three structural roles.
In a \textbf{chain} ($V_{m-1} \to V_m \to V_{m+1}$), $V_m$ is a causal
intermediary through which influence is transmitted.  In a \textbf{fork}
($V_{m-1} \leftarrow V_m \to V_{m+1}$), $V_m$ is a common cause that
creates confounding between the two endpoints.  In a \textbf{collider}
($V_{m-1} \to V_m \leftarrow V_{m+1}$), both arrows point \emph{into}
$V_m$; this is the crucial asymmetric case whose behavior under
conditioning is the opposite of the other two, as
Figure~\ref{w2:fig:chain-fork-collider} shows.

\begin{figure}[t]
\centering
\resizebox{\textwidth}{!}{%
\begin{tikzpicture}[>=stealth, node distance=1.6cm and 1.8cm]
  %% -------- Chain --------
  \node[node] (A1) at (0,0)   {$A$};
  \node[node] (M1) at (1.8,0) {$M$};
  \node[node] (B1) at (3.6,0) {$B$};
  \draw[edge] (A1) -- (M1);
  \draw[edge] (M1) -- (B1);
  \node[font=\small\bfseries\color{isublue}] at (1.8, 0.9) {Chain};
  \node[font=\small\color{darkgrey}] at (1.8,-0.9)
    {$A \nindep B$
     \quad but \quad
     $A \indep B \mid M$};
  %% -------- Fork --------
  \node[node] (A2) at (6.2,0) {$A$};
  \node[node] (M2) at (8.0,0) {$M$};
  \node[node] (B2) at (9.8,0) {$B$};
  \draw[edge] (M2) -- (A2);
  \draw[edge] (M2) -- (B2);
  \node[font=\small\bfseries\color{isublue}] at (8.0, 0.9) {Fork};
  \node[font=\small\color{darkgrey}] at (8.0,-0.9)
    {$A \nindep B$
     \quad but \quad
     $A \indep B \mid M$};
  %% -------- Collider --------
  \node[node] (A3) at (12.4,0) {$A$};
  \node[node] (M3) at (14.2,0) {$M$};
  \node[node] (B3) at (16.0,0) {$B$};
  \draw[edge] (A3) -- (M3);
  \draw[edge] (B3) -- (M3);
  \node[font=\small\bfseries\color{isublue}] at (14.2, 0.9) {Collider};
  \node[font=\small\color{darkgrey}] at (14.2,-0.9)
    {$A \indep B$
     \quad but \quad
     $A \nindep B \mid M$};
  %% -------- Vertical dividers --------
  \draw[dotted, color=accent!40, thin] (5.0, 1.4) -- (5.0,-1.4);
  \draw[dotted, color=accent!40, thin] (11.0, 1.4) -- (11.0,-1.4);
\end{tikzpicture}%
}
\caption{The three fundamental path structures in a DAG, with their default
independence status and the effect of conditioning on the middle node $M$.
Chains and forks are \emph{open by default} and are blocked by conditioning
on $M$.  Colliders are \emph{closed by default} and are opened by
conditioning on $M$ (or any descendant of $M$).}
\label{w2:fig:chain-fork-collider}
\end{figure}

\subsection{Blocking Rules}

Conditioning on a node affects the paths it lies on differently depending on
its structural role:

\begin{center}
\renewcommand{\arraystretch}{1.35}
\begin{tabular}{@{}llp{5.8cm}p{4.0cm}@{}}
\toprule
\textbf{Structure} & \textbf{Pattern} & \textbf{Effect of conditioning} & \textbf{Key intuition} \\
\midrule
Chain    & $A \to M \to B$    & Blocks the path & Causal transmission \\
Fork     & $A \leftarrow M \to B$ & Blocks the path & Common cause \\
Collider & $A \to M \leftarrow B$ & \textbf{Opens} the path (conditioning on $M$ or a descendant of $M$) & Common effect \\
\bottomrule
\end{tabular}
\end{center}

\medskip
The critical asymmetry --- colliders are \emph{closed by default} and
\emph{opened} by conditioning, while chains and forks are \emph{open by
default} and \emph{closed} by conditioning --- is what makes collider bias
so easy to overlook in practice.  The following three examples anchor each
rule to a concrete story.

\medskip
\noindent\textbf{Chain: Smoking $\to$ Tar $\to$ Cancer.}
Let $A$ = smoking, $M$ = tar deposits in the lungs, $B$ = lung cancer.
The path $A \to M \to B$ is a chain: smoking causes tar accumulation, which
in turn causes cancer.
Marginally, heavy smokers have elevated cancer rates --- the path is open.
But if we condition on tar level (e.g., restrict to patients with the same
measured tar burden), smoking no longer predicts cancer within that stratum:
once the tar level is fixed, the causal channel is fully accounted for.
Conditioning on the intermediary $M$ \emph{blocks} the path.

\medskip
\noindent\textbf{Fork: Poverty $\to$ Poor Diet; Poverty $\to$ Lack of Exercise.}
Let $M$ = poverty, $A$ = poor diet, $B$ = lack of exercise.
The path $A \leftarrow M \to B$ is a fork: poverty is a common cause of
both outcomes.
Unconditionally, diet quality and exercise are correlated --- not
because one causes the other, but because both are driven by poverty.
Conditioning on income level removes this spurious association: among people
at the same income, the diet--exercise correlation disappears.
Conditioning on the common cause $M$ \emph{blocks} the path.

\medskip
\noindent\textbf{Collider: Accident $\to$ hospitalization $\leftarrow$ Cancer.}
Let $A$ = traffic accident, $B$ = cancer, $M$ = hospitalization.
Both accidents and cancer independently cause hospitalization, so the path
$A \to M \leftarrow B$ is a collider at $M$.
In the general population, having a traffic accident tells you nothing about
your cancer risk: $A \indep B$.
But among hospitalised patients --- conditioning on $M$ --- the two become
negatively associated.
If we learn that a hospitalised patient was \emph{not} in a traffic accident,
cancer or another serious illness becomes more likely as an explanation for
the admission.
Conditioning on the common effect $M$ \emph{opens} the path.
This is Berkson's bias \citep{berkson1946limitations}.

\medskip
The three examples above share a single underlying logic, summarized in
the following reference box.

\begin{tcolorbox}[defstyle, title={Three-Node Motifs at a Glance}]
The three basic path structures behave differently under conditioning.
In a \emph{chain} $A \to M \to B$, the middle node $M$ transmits
association, so conditioning on $M$ blocks the path.
In a \emph{fork} $A \leftarrow M \to B$, the middle node $M$ is a common
cause, so conditioning on $M$ removes the spurious association.
In a \emph{collider} $A \to M \leftarrow B$, the path is blocked by
default, but conditioning on $M$ --- or on any descendant of $M$ ---
opens the path and induces association.
These three motifs are the local building blocks of d-separation in
arbitrary DAGs.
\end{tcolorbox}

\subsection{The d-Separation Criterion}

Before stating the formal criterion, it is useful to summarize the
intuition: a path stays active unless we block a non-collider on it,
or unless it contains a collider that has not been conditioned on (and
has no conditioned-on descendant).
The definition below simply turns this intuition into a rule that applies
to any path in any DAG.

\begin{defn}{d-Blocking and d-Separation}
\label{w2:def:dsep}
A path $\pi$ is \emph{d-blocked} by a set $\mathbf{S}$ if:
\begin{enumerate}
  \item $\pi$ contains a chain $A \to M \to B$ or fork $A \leftarrow M \to B$
        with $M \in \mathbf{S}$; or
  \item $\pi$ contains a collider $A \to C \leftarrow B$ with $C \notin
        \mathbf{S}$ and no descendant of $C$ is in $\mathbf{S}$.
\end{enumerate}
Nodes $X$ and $Y$ are \emph{d-separated} by $\mathbf{S}$, written
$(X \indep Y \mid \mathbf{S})_{\Gcal}$, if every path between
$X$ and $Y$ in $\Gcal$ is d-blocked by $\mathbf{S}$.
\end{defn}

\begin{example}{Applying Definition~\ref{w2:def:dsep} to the Three Toy Settings}
\label{w2:ex:threesettings}
We revisit the three examples from Section~2.2 and verify each blocking
claim directly using Definition~\ref{w2:def:dsep}.

\medskip
\noindent\textbf{(i) Chain --- Smoking $\to$ Tar $\to$ Cancer.}
Let $A$ = smoking, $M$ = tar, $B$ = cancer.
The only path between $A$ and $B$ is the chain $A \to M \to B$.

\begin{itemize}
  \item $\mathbf{S} = \varnothing$: the path contains no conditioned-on
        non-collider, so clause~(1) does not apply.
        The collider clause~(2) is also irrelevant (there is no collider).
        The path is \emph{not} d-blocked.
        $\Rightarrow A \nindep B$ (smokers have higher cancer rates).
  \item $\mathbf{S} = \{M\}$: the chain node $M \in \mathbf{S}$,
        so clause~(1) applies and the path is d-blocked.
        $\Rightarrow A \indep B \mid M$ (at fixed tar level,
        smoking no longer predicts cancer).
\end{itemize}

\medskip
\noindent\textbf{(ii) Fork --- Poverty $\to$ Poor Diet; Poverty $\to$ Lack of Exercise.}
Let $M$ = poverty, $A$ = poor diet, $B$ = lack of exercise.
The only path is the fork $A \leftarrow M \to B$.

\begin{itemize}
  \item $\mathbf{S} = \varnothing$: no non-collider on the path is in
        $\mathbf{S}$, so the path is not d-blocked.
        $\Rightarrow A \nindep B$ (poor diet and lack of
        exercise are correlated through shared poverty).
  \item $\mathbf{S} = \{M\}$: the fork node $M \in \mathbf{S}$, so
        clause~(1) applies and the path is d-blocked.
        $\Rightarrow A \indep B \mid M$ (at fixed income level,
        the diet--exercise association disappears).
\end{itemize}

\medskip
\noindent\textbf{(iii) Collider --- Accident $\to$ hospitalization $\leftarrow$ Cancer.}
Let $A$ = accident, $M$ = hospitalization, $B$ = cancer.
The only path is the collider $A \to M \leftarrow B$.

\begin{itemize}
  \item $\mathbf{S} = \varnothing$: the path contains collider $M$
        with $M \notin \mathbf{S}$ and no descendant of $M$ in
        $\mathbf{S}$, so clause~(2) applies and the path is d-blocked.
        $\Rightarrow A \indep B$ (accidents and cancer are
        independent in the general population).
  \item $\mathbf{S} = \{M\}$: now $M \in \mathbf{S}$, so clause~(2) no
        longer blocks the path.  No other clause applies, so the path is
        \emph{open}.
        $\Rightarrow A \nindep B \mid M$ (among hospitalised
        patients, ruling out an accident raises the probability of cancer ---
        Berkson's bias).
\end{itemize}
\end{example}

Example~\ref{w2:ex:threesettings} showed that d-separation is a purely graphical
calculation.  Before stating the theorem that gives it statistical force,
we verify the claim for the chain directly, showing that the graphical
separation statement is an ordinary conditional-independence identity in
the distribution.

\begin{example}{A Direct Probability Proof for the Chain}
\label{w2:ex:chainproof}
Consider the chain $X_1 \to X_2 \to X_3$.  By the Markov factorization,
\[
  p(x_1, x_2, x_3) = p(x_1)\,p(x_2 \mid x_1)\,p(x_3 \mid x_2).
\]
Conditioning on $X_2 = x_2$:
\[
  p(x_1, x_3 \mid x_2)
  = \frac{p(x_1, x_2, x_3)}{p(x_2)}
  = \frac{p(x_1)\,p(x_2 \mid x_1)\,p(x_3 \mid x_2)}{p(x_2)}.
\]
Rearranging,
\[
  p(x_1, x_3 \mid x_2)
  = \frac{p(x_1)\,p(x_2 \mid x_1)}{p(x_2)}\,p(x_3 \mid x_2)
  = p(x_1 \mid x_2)\,p(x_3 \mid x_2).
\]
Hence $X_1 \indep X_3 \mid X_2$.
This derivation shows concretely how a graphical separation statement
becomes an ordinary conditional-independence identity in the distribution.
The corresponding fork case $X_1 \leftarrow X_2 \to X_3$ can be verified
similarly and is left as an exercise.
\end{example}

The following theorem is what gives d-separation its statistical force: it
guarantees that every d-separation statement in the graph corresponds to a
genuine conditional independence in the joint distribution.

\begin{thm}{Verma \& Pearl, 1988; \citep{pearl2009causality}, Theorem~1.2.4}
\label{w2:thm:dsep}
If $(X \indep Y \mid \mathbf{S})_{\Gcal}$, then
$X \indep Y \mid \mathbf{S}$ in every distribution that is Markov
with respect to $\Gcal$.
\end{thm}

\begin{remark}
\textbf{Soundness, not converse.}
Theorem~\ref{w2:thm:dsep} states that d-separation implies conditional
independence for every distribution Markov with respect to $\Gcal$.
The converse need not hold without additional assumptions such as
\emph{faithfulness}: a conditional independence may occur in the data
because of special parameter values even when the corresponding nodes
are not d-separated in the graph.
\end{remark}

Example~\ref{w2:ex:chainproof} illustrates in the simplest chain case how
the Markov factorization turns a graphical d-separation statement into an
ordinary conditional-independence identity in the distribution.

\subsection{Practical Ways to Check d-Separation}

The formal definition of d-separation can appear abstract when applied to
larger graphs.  Two complementary algorithmic viewpoints are useful in
practice, both equivalent to the definition.

\paragraph{1. Bayes-Ball Intuition.}
Imagine releasing a ball from $X$ and asking whether it can reach $Y$,
given that nodes in $\mathbf{S}$ are observed (shaded).
The ball follows four local rules:
\begin{itemize}
  \item At a chain or fork node \emph{not} in $\mathbf{S}$: ball passes
        through.
  \item At a chain or fork node in $\mathbf{S}$: ball stops.
  \item At a collider not in $\mathbf{S}$ (and no descendant in
        $\mathbf{S}$): ball stops.
  \item At a collider in $\mathbf{S}$ (or with a descendant in
        $\mathbf{S}$): ball passes through.
\end{itemize}
If the ball can reach $Y$ from $X$, then $X \nindep Y \mid
\mathbf{S}$.  If every path is stopped, then $X \indep Y \mid
\mathbf{S}$.  The algorithm is formalised in \citet{shachter1998bayesball}.

\paragraph{2. Moral Graph Transformation.}
There is also a graph-transformation approach, which connects DAG models
to undirected graphical models (Markov random fields).
\begin{enumerate}
  \item Restrict the DAG to the \emph{ancestral set}
        $\An(X \cup Y \cup \mathbf{S})$: discard all nodes that
        are not ancestors of $X$, $Y$, or any element of $\mathbf{S}$.
  \item \emph{Moralise} the graph: for every collider $A \to C \leftarrow
        B$ in the ancestral subgraph, add an undirected edge $A - B$
        (``marry the parents'').
  \item Remove all edge directions to obtain an undirected graph.
  \item Delete all nodes in $\mathbf{S}$ (and their incident edges).
\end{enumerate}
If $X$ and $Y$ are disconnected in the resulting graph, then
$(X \indep Y \mid \mathbf{S})_{\Gcal}$.
The moralization step is what makes the collider rule visible in the
undirected representation: without adding the $A$--$B$ edge, the path
through $C$ would appear disconnected even when $C$ is conditioned upon,
which would give the wrong answer.  A full treatment is in
\citet[][Ch.~3]{lauritzen1996graphical}.

\medskip
These two viewpoints --- Bayes-ball propagation and moral graph separation
--- are equivalent algorithmic ways of evaluating the d-separation criterion.
Students are encouraged to use whichever they find most natural, and to
cross-check with the other when in doubt.

\begin{example}{Checking d-Separation in a Four-Node DAG}
\label{w2:ex:fournode}
Consider the DAG with edges $Z \to T$, $U \to T$, $T \to Y$, $U \to Y$.

\begin{center}
\begin{tikzpicture}[node distance=1.8cm]
  \node[node] (Z) {$Z$};
  \node[node, right of=Z] (T) {$T$};
  \node[node, right of=T] (Y) {$Y$};
  \node[node, below=1.0cm of T] (U) {$U$};
  \draw[edge] (Z) -- (T);
  \draw[edge] (T) -- (Y);
  \draw[edge] (U) -- (T);
  \draw[edge] (U) -- (Y);
\end{tikzpicture}
\end{center}

We apply both methods to two queries.

\medskip
\noindent\textbf{Query 1.} Is $(Z \indep U)_{\Gcal}$,
i.e.\ $\mathbf{S} = \varnothing$?

\medskip
\noindent\emph{Bayes-ball.}
The only path from $Z$ to $U$ is $Z \to T \leftarrow U$.
Node $T$ is a collider with $T \notin \mathbf{S}$ and no descendant of $T$
in $\mathbf{S}$, so the ball is stopped at $T$.
No ball can reach $U$, hence $(Z \indep U)_{\Gcal}$.

\medskip
\noindent\emph{Moral graph.}
\begin{enumerate}
  \item \emph{Ancestral set} of $\{Z, U\} \cup \mathbf{S} = \{Z, U\}$:
        neither $Z$ nor $U$ has any parents, so
        $\An(\{Z,U\}) = \{Z, U\}$.
        Nodes $T$ and $Y$ are discarded.
  \item \emph{Moralise} the subgraph $\{Z, U\}$: it has no edges and no
        colliders, so nothing is added.
  \item \emph{Undirected graph}: isolated nodes $Z$ and $U$.
  \item \emph{Delete} $\mathbf{S} = \varnothing$: nothing changes.
\end{enumerate}
$Z$ and $U$ are disconnected $\Rightarrow$ $(Z \indep U)_{\Gcal}$.
Note that the key step is restricting to the ancestral set: this removes $T$
from the graph before moralization, so the collider $T$ never has a chance
to create a moral edge between $Z$ and $U$.

\medskip
\noindent\textbf{Query 2.} Is $(Z \indep U \mid T)_{\Gcal}$,
i.e.\ $\mathbf{S} = \{T\}$?

\medskip
\noindent\emph{Bayes-ball.}
Same path $Z \to T \leftarrow U$.
Now $T \in \mathbf{S}$: the collider is observed, so the ball passes through $T$.
The path is open, hence $Z \nindep U \mid T$.

\medskip
\noindent\emph{Moral graph.}
\begin{enumerate}
  \item \emph{Ancestral set} of $\{Z, U\} \cup \mathbf{S} = \{Z, U, T\}$:
        parents of $T$ are $Z$ and $U$, both already in the set.
        $\An(\{Z,U,T\}) = \{Z, U, T\}$; node $Y$ is discarded.
  \item \emph{Moralise}: the subgraph contains the collider $Z \to T
        \leftarrow U$, so add the moral edge $Z - U$.
  \item \emph{Undirected graph}: edges $Z - T$, $U - T$, $Z - U$.
  \item \emph{Delete} $\mathbf{S} = \{T\}$ (and its incident edges $Z-T$, $U-T$).
        Remaining graph: single edge $Z - U$.
\end{enumerate}
$Z$ and $U$ are connected $\Rightarrow$ $Z \nindep U \mid T$.
The moralization step is decisive: it adds the edge $Z - U$ \emph{before}
$T$ is deleted, ensuring the dependence induced by conditioning on the
collider is visible in the undirected graph.

\medskip
Comparing the two queries: $Z$ and $U$ are marginally independent (the
instrument is exogenous) but become dependent once we condition on $T$.
This is Berkson's bias, treated in full in Section~\ref{w2:sec:ivdag}.
\end{example}

\begin{example}{Practice DAG --- Eight-Node Graph}
\label{w2:ex:practicedag}
Consider the DAG below with nodes $W$, $X_1$, $T$, $M_1$, $M_2$, $Y$,
$C$, and $D$.

\begin{center}
\begin{tikzpicture}[node distance=1.8cm]
  \node[node] (W)  at (1.8,  2.0) {$W$};
  \node[node] (X1) at (0.0,  0.0) {$X_1$};
  \node[node] (T)  at (1.8,  0.0) {$T$};
  \node[node] (M1) at (3.6,  0.0) {$M_1$};
  \node[node] (Y)  at (5.4,  0.0) {$Y$};
  \node[node] (M2) at (3.6, -1.6) {$M_2$};
  \node[node] (C)  at (1.8, -1.6) {$C$};
  \node[node] (D)  at (1.8, -3.2) {$D$};
  % edges
  \draw[edge] (W)  -- (X1);
  \draw[edge] (W)  -- (T);
  \draw[edge] (W)  -- (Y);
  \draw[edge] (T)  -- (M1);
  \draw[edge] (M1) -- (Y);
  \draw[edge] (M1) -- (M2);
  \draw[edge] (X1) -- (C);
  \draw[edge] (M2) -- (C);
  \draw[edge] (C)  -- (D);
\end{tikzpicture}
\end{center}

For each query below, state whether the independence holds and identify
every relevant path and its blocking status.

\medskip
\noindent\textbf{(1)} Is $X_1 \indep Y$?

\noindent\emph{Answer: No.}
There are two open paths from $X_1$ to $Y$:
$X_1 \leftarrow W \to Y$ (fork at $W$, unblocked)
and $X_1 \leftarrow W \to T \to M_1 \to Y$ (fork at $W$, then chain; also open).
The path $X_1 \to C \leftarrow M_2$ ends at a collider $C$ with $C \notin \varnothing$,
so it is blocked. One open path suffices: $X_1 \nindep Y$.

\medskip
\noindent\textbf{(2)} Is $X_1 \indep Y \mid W$?

\noindent\emph{Answer: Yes.}
Both open paths above pass through the fork $W$; conditioning on $W$ blocks them.
The remaining path $X_1 \to C \leftarrow M_2 \leftarrow M_1 \leftarrow T \leftarrow W \to Y$
has a collider at $C$ with $C \notin \{W\}$ (and no descendant of $C$ in $\{W\}$),
so it remains blocked.
All paths blocked $\Rightarrow X_1 \indep Y \mid W$.

\medskip
\noindent\textbf{(3)} Is $T \indep M_2 \mid M_1$?

\noindent\emph{Answer: Yes.}
The direct path $T \to M_1 \to M_2$ is a chain with $M_1 \in \{M_1\}$ --- blocked.
The path $T \leftarrow W \to X_1 \to C \leftarrow M_2$ has a collider at $C$
with $C \notin \{M_1\}$ --- blocked.
The path $T \leftarrow W \to Y \leftarrow M_1 \to M_2$ has a collider at $Y$
($W \to Y \leftarrow M_1$) with $Y \notin \{M_1\}$ --- blocked.
All paths blocked $\Rightarrow T \indep M_2 \mid M_1$.

\medskip
\noindent\textbf{(4)} Does conditioning on $C$ create collider bias?

\noindent\emph{Answer: Yes.}
$C$ is a collider on the path $X_1 \to C \leftarrow M_2$.
Marginally, $X_1 \indep M_2$ (that path is blocked by the collider at $C$).
Conditioning on $C$ opens the path, inducing a spurious association
between $X_1$ and $M_2$ that does not reflect any direct causal link.

\medskip
\noindent\textbf{(5)} Does conditioning on $D$ open the path $X_1 \to C \leftarrow M_2$?

\noindent\emph{Answer: Yes.}
$D$ is a descendant of $C$.  By clause~(2) of Definition~\ref{w2:def:dsep},
a collider path is unblocked whenever the collider \emph{or any of its
descendants} is in the conditioning set.
Since $D \in \{D\}$, the collider at $C$ is activated, opening the path
$X_1 \to C \leftarrow M_2$ even though $C$ itself is not conditioned on.
\end{example}

% ==============================================================
\section{Collider Bias and the IV DAG}
\label{w2:sec:ivdag}
% ==============================================================

\subsection{Berkson's Bias}

\begin{defn}{Collider Bias}
\label{w2:def:colliderbias}
\emph{Collider bias} is the spurious association induced between two
variables when one conditions on their common effect (a collider), or
on a descendant of that collider.
In a path $A \to C \leftarrow B$, the variables $A$ and $B$ may be
marginally independent, but conditioning on $C$ --- or on any descendant
of $C$ --- can make them statistically dependent.
\end{defn}

\begin{warn}{Collider bias vs.\ confounding}
Unlike confounding, which is \emph{removed} by conditioning on a common
cause, collider bias is \emph{created} by conditioning on a common
effect.
This is why conditioning on more variables is not always safer in causal
analysis: adding a collider or a descendant of a collider to the
adjustment set introduces bias rather than reducing it.
\end{warn}

\begin{example}{Collider Bias in the IV DAG}
In the IV DAG, $T$ is a collider on the path $Z \to T \leftarrow U$.
Marginally, $Z \indep U$ (the instrument is exogenous). However,
\[
  Z \indep U \qquad \text{but} \qquad
  Z \nindep U \mid T
\]
(established graphically in Example~\ref{w2:ex:fournode}).
Conditioning on $T$ --- restricting to the treated subpopulation --- makes
the instrument $Z$ appear associated with $U$,
invalidating the IV argument within that stratum.
\end{example}

\begin{example}{Collider Bias through Selection: Talent and Wealth}
\label{w2:ex:admissions}
A similar phenomenon arises whenever selection depends on two otherwise
unrelated variables.
Suppose both talent ($A$) and wealth ($B$) increase the probability of
admission to an elite school, so admission ($S$) is a collider:
$A \to S \leftarrow B$.

\begin{center}
\begin{tikzpicture}[node distance=2.0cm]
  \node[node] (A) {$A$};
  \node[node, right of=A] (S) {$S$};
  \node[node, right of=S] (B) {$B$};
  \draw[edge] (A) -- (S);
  \draw[edge] (B) -- (S);
\end{tikzpicture}
\end{center}

In the general population, talent and wealth may be independent: the
path $A \to S \leftarrow B$ is blocked by the collider $S$, so
$A \indep B$.
But among admitted students --- conditioning on $S$ --- lower talent
makes higher wealth more likely, and lower wealth makes higher talent
more likely, because both variables contribute to the same selection
event.
Conditioning on admission therefore induces association between variables
that were not associated in the population.
This is again collider bias: admission is a common effect of talent and
wealth, and restricting to admitted students is precisely conditioning on
that common effect.
\end{example}

\subsection{Full d-Separation Analysis of the IV DAG}

We work through the IV DAG with edges $Z \to T$, $T \to Y$, $U \to T$,
$U \to Y$.  This is the same graph as in Example~\ref{w2:ex:fournode}, but now
$U$ is treated as \emph{unobserved}.  That assumption is precisely what earns
the name ``IV DAG'': because $U$ cannot be conditioned on, the back-door path
$T \leftarrow U \to Y$ cannot be blocked by adjustment, and the instrument
$Z$ becomes the only route to identification.

Before listing the paths, it is worth classifying them by \emph{type},
since students often conflate causal and associational dependence.

\begin{center}
\renewcommand{\arraystretch}{1.35}
\begin{tabular}{@{}llll@{}}
\toprule
\textbf{Endpoints} & \textbf{Path} & \textbf{Type} & \textbf{Why} \\
\midrule
$Z \leftrightarrow Y$ & $Z \to T \to Y$ & \emph{Causal} & Directed from $Z$ to $Y$; carries the IV signal \\
$Z \leftrightarrow Y$ & $Z \to T \leftarrow U \to Y$ & \emph{Associational} & Collider at $T$; activates back-door via $U$ \\
$Z \leftrightarrow U$ & $Z \to T \leftarrow U$ & \emph{Associational} & Collider at $T$; dormant unless $T$ conditioned on \\
\bottomrule
\end{tabular}
\end{center}

\medskip
\noindent The key distinction: a \emph{causal path} follows all edges in
their directed orientation from source to target.
An \emph{associational path} travels against at least one arrow, or passes
through a collider, and therefore does not represent a mechanism by which
$Z$ produces $Y$ --- it can nevertheless generate a statistical association
if the collider is conditioned upon.
With this in mind, the four questions become much easier to diagnose.

\medskip
\noindent\textbf{Question 1.} Is $Z \indep Y$?

\noindent There are two paths between $Z$ and $Y$.
The directed path $Z \to T \to Y$ is open, since it is a chain and we
are not conditioning on $T$.
The path $Z \to T \leftarrow U \to Y$ contains a collider at $T$.
Since we are not conditioning on $T$ or any descendant of $T$, that path
is blocked.
Because the causal path $Z \to T \to Y$ remains open, we conclude that
$Z \nindep Y$.

\medskip
\noindent\textbf{Question 2.} Is $Z \indep Y \mid T$?

\noindent The causal path $Z \to T \to Y$ is blocked because conditioning
on $T$ blocks the chain.
The path $Z \to T \leftarrow U \to Y$ contains a collider at $T$, and
because we are conditioning on $T$, that collider is opened.
$\Rightarrow Z \nindep Y \mid T$.

\medskip
\noindent\textbf{Question 3.} Is $Z \indep Y \mid \{T, U\}$?

\noindent The causal path $Z \to T \to Y$ is blocked because conditioning
on $T$ blocks the chain.
The path $Z \to T \leftarrow U \to Y$ contains a collider at $T$, and
because we are conditioning on $T$, that collider is opened.
However, the same path then passes through $U$, and because we are also
conditioning on $U$, the path is blocked there.
Hence both the causal path and the back-door path are blocked, so
$Z \indep Y \mid \{T, U\}$.

\medskip
\noindent\textbf{Question 4.} Is $Z \indep U$?

\noindent The only path between $Z$ and $U$ is $Z \to T \leftarrow U$.
This path contains a collider at $T$.
Since we are not conditioning on $T$ or on any descendant of $T$, the
collider remains closed, so the path is blocked.
Hence $Z \indep U$.

\medskip
Taken together, the four questions give a complete graphical account of the
three IV assumptions and one critical warning.
Question~1 establishes \emph{relevance}: $Z$ has an open causal path to $Y$
through $T$, so the instrument is not inert.
Question~4 establishes \emph{exogeneity}: the only path from $Z$ to the
confounder $U$ is a dormant collider path, so the instrument is
independent of unmeasured confounding.
Question~3 establishes the \emph{exclusion restriction}: once both $T$ and
$U$ are held fixed, $Z$ carries no residual information about $Y$, meaning
the instrument affects the outcome only through the treatment channel.
Note, however, that this conditioning requires observing $U$ --- which is
unavailable by assumption. The IV strategy exploits the exclusion restriction
indirectly, through the moment condition $\E[\varepsilon \cdot Z] = 0$ derived
in Chapter~7, without ever observing $U$.
Finally, Question~2 is the warning: conditioning on $T$ alone --- as a naive
analyst might do --- simultaneously closes the causal channel and opens the
confounded channel, producing a spurious association that is worse than no
adjustment at all.

It is important to recognize that the IV DAG encodes the substantive
assumptions behind instrument validity graphically --- relevance, exogeneity,
and exclusion --- but does not make them testable in any strong sense.
Some observed-data implications of the IV DAG may be checked for
compatibility with data, but the core IV assumptions are not generally
fully testable from observational data alone.

% ==============================================================
\section{The Markov Property and factorization}
\label{w2:sec:markov}
% ==============================================================

Up to this point, we have used DAGs qualitatively, to decide which paths
are open or blocked.
We now connect the graph to probability algebra: the same parent structure
that governs d-separation also determines how the joint distribution
factorizes.

Without structural assumptions, any joint distribution can always be
written by repeated conditioning, but that generic representation is
often too high-dimensional to reveal much structure.
A DAG becomes statistically meaningful because, together with the Markov
property, it replaces the generic factorization by a sparse one involving
only the parents of each node.
This is the sense in which a graphical model is not merely a picture: it
imposes testable probabilistic structure on the joint distribution.

The following proposition makes this concrete for the simplest possible
case.  The proof was carried out in Example~\ref{w2:ex:chainproof}; we
state the result formally here for reference.

\begin{prop}{Conditional independence in the three-node chain}
\label{w2:prop:chain}
Let $X_1 \to X_2 \to X_3$ be a chain with Markov factorization
$p(x_1, x_2, x_3) = p(x_1)\,p(x_2 \mid x_1)\,p(x_3 \mid x_2)$.
Then $X_1 \indep X_3 \mid X_2$.
The analogous result for the fork $X_1 \leftarrow X_2 \to X_3$ is left
as an exercise.
\end{prop}

\begin{defn}{Local Markov Property}
A distribution $P$ satisfies the \emph{local Markov property} with respect
to $\Gcal$ if, for every node $V_i \in \mathcal{V}$:
\[
  V_i \indep \bigl(\Nd(V_i) \setminus \Pa(V_i)\bigr) \mid \Pa(V_i).
\]
That is, once we condition on the direct causes of a node, that node is
independent of all variables that are neither its descendants nor its parents.
\end{defn}

\begin{remark}
The \emph{global Markov property} is the statement that every
d-separation in $\Gcal$ implies a conditional independence in $P$
--- precisely the content of Theorem~\ref{w2:thm:dsep}.
For DAGs, the local and global Markov properties are equivalent
\citep[][Proposition~3.27]{lauritzen1996graphical}: the local property
implies the global one via the Markov factorization, which is what is
actually used in identification arguments.
\end{remark}

\begin{remark}
In a causal DAG, an arrow is typically drawn only when a direct
dependence-generating relation is believed to be present.
For that reason, adjacent nodes are usually expected to be statistically
dependent unless special parameter cancellations occur.
We do not formalize minimality or faithfulness here, but this intuition
explains why edges are not included gratuitously: every arrow in the graph
represents a substantive scientific claim.
\end{remark}

An immediate consequence is the \emph{Markov factorization}:
\begin{equation}
  p(v_1, \dots, v_k) \;=\; \prod_{i=1}^{k} p\!\left(v_i \mid \Pa(v_i)\right).
  \label{w2:eq:markov}
\end{equation}
This factorization is the bridge between causal structure and probability:
knowing the parents of each node is sufficient to write the full joint density.

\begin{example}{Markov factorization in the Education Example}
For the DAG with edges $N \to E$, $B \to E$, $B \to Y$, $E \to Y$:
\[
  p(n, b, e, y) = p(n)\,p(b)\,p(e \mid n, b)\,p(y \mid e, b).
\]
A regression of $Y$ on $E$ alone is confounded by $B$. Conditioning on $B$
blocks the back-door path $E \leftarrow B \to Y$ and identifies the causal
effect $P(y \mid \doop(E{=}e))$ via the back-door formula (Chapter~3).
\end{example}

\begin{example}{Markov factorization in the IV DAG}
Including the latent $U$:
\[
  p(z, t, y, u) = p(z)\,p(u)\,p(t \mid z, u)\,p(y \mid t, u).
\]
Since $U$ is unobserved, the observable factorization is obtained by
marginalising:
\[
  p(z, t, y) = \int p(z)\,p(u)\,p(t \mid z, u)\,p(y \mid t, u)\,\mathrm{d}u.
\]
This cannot be simplified to $p(z)\,p(t\mid z)\,p(y\mid t)$ when $U$ is a
common cause of $T$ and $Y$: this is precisely the endogeneity problem.
\end{example}

% ==============================================================
\section{Worked Example: The Education--Earnings DAG}
\label{w2:sec:education}
% ==============================================================

This section is the template for how we will use DAGs throughout the course.
Starting from a single causal graph, we move step by step through the full
pipeline: first specify the DAG, then read off the factorization, then use
d-separation to identify the implied conditional independences, and finally
interpret the result in causal terms.
Later chapters will follow the same pattern, but with richer identification
questions.

\medskip
\noindent\textbf{Practice: d-separation workflow.}
Before working through the example below, return to
Example~\ref{w2:ex:practicedag} and rework each of the five queries from
scratch, without looking at the answers.
For each query, follow the same three steps:
(1)~list every path between the two nodes of interest;
(2)~classify every intermediate node on each path as a chain, fork, or
collider; and
(3)~determine which paths are blocked or open after conditioning on the
given set.
This is exactly the procedure that d-separation always requires, in graphs
of any size.

\medskip
\noindent\textbf{The causal story.}
Education ($E$) affects earnings ($Y$); family background ($B$) is a
common cause of both education and earnings; neighborhood ($N$) affects
education but has no direct effect on earnings.  All four variables are
observed, so the graph contains no latent confounders.

\begin{center}
\begin{tikzpicture}[node distance=1.8cm]
  \node[node] (N) {$N$};
  \node[node, right of=N] (E) {$E$};
  \node[node, right of=E] (Y) {$Y$};
  \node[node, below=1.0cm of E] (B) {$B$};
  \draw[edge] (N) -- (E);
  \draw[edge] (E) -- (Y);
  \draw[edge] (B) -- (E);
  \draw[edge] (B) -- (Y);
\end{tikzpicture}
\end{center}

\medskip
\noindent\textbf{Step 1 --- Markov factorization.}
The parents are $\Pa(N) = \Pa(B) = \varnothing$,
$\Pa(E) = \{N, B\}$, and $\Pa(Y) = \{E, B\}$.
The Markov factorization is therefore
\[
  p(n, b, e, y) \;=\; p(n)\,p(b)\,p(e \mid n, b)\,p(y \mid e, b).
\]
Every factor on the right-hand side involves only observed variables, so
in principle each term is estimable from data.
This is the starting point for identification: the full joint distribution
is expressed in terms of estimable quantities.

\medskip
\noindent\textbf{Step 2 --- d-Separation.}
There are two paths between $N$ and $Y$:
\begin{itemize}
  \item Path 1: $N \to E \to Y$ (a chain through $E$).
  \item Path 2: $N \to E \leftarrow B \to Y$ (a collider at $E$, followed
        by the fork leg $B \to Y$).
\end{itemize}
We examine four queries.

\smallskip
\emph{Query (i): Is $(N \indep Y)_{\Gcal}$?}
Path~1 is a chain with nothing conditioned on, so it is open.
Path~2 has a collider at $E$, and since $E$ is not conditioned on, that
path is blocked.
Because Path~1 remains open, we conclude that $N \nindep Y$.

\smallskip
\emph{Query (ii): Is $(N \indep Y \mid E)_{\Gcal}$?}
Path~1 is a chain, and conditioning on $E$ blocks it.
Path~2 has a collider at $E$, and conditioning on $E$ opens it.
The activated path $N \to E \leftarrow B \to Y$ continues through $B$,
which is not conditioned on, so the path is open.
Hence $N \nindep Y \mid E$.
This is a collider trap: conditioning on $E$ opens the path
$N \to E \leftarrow B \to Y$, thereby inducing a spurious association
between $N$ and $Y$ through $B$.

\smallskip
\emph{Query (iii): Is $(N \indep Y \mid \{E, B\})_{\Gcal}$?}
Path~1 is blocked by conditioning on $E$.
Path~2 is opened at the collider $E$, but then blocked at $B$ because
$B$ is also conditioned on.
Both paths are blocked: $(N \indep Y \mid E, B)_{\Gcal}$.

\smallskip
\emph{Query (iv): Is $(N \indep B)_{\Gcal}$?}
The only path is $N \to E \leftarrow B$, which has a collider at $E$.
Since $E$ is not conditioned on, the path is blocked.
Therefore $N \indep B$.

\medskip
\noindent\textbf{Step 3 --- Conditional Independence.}
We now move from the graphical analysis to the probabilistic implications.
Up to this point, we have reasoned in graphical terms by identifying open
and blocked paths.
The next step is to translate those graphical conclusions into conditional
independence statements in the distribution, using the Markov property.
By Theorem~\ref{w2:thm:dsep}, the d-separation results imply
\[
  N \indep B, \qquad
  N \indep Y \mid E, B, \qquad
  N \nindep Y, \qquad
  N \nindep Y \mid E.
\]
The first two are independence statements; the last two are dependence
statements.
Three of the four are, in principle, assessable from observed data and
can be used to probe whether the graph is compatible with the data.
Failure of such implications can falsify the graph, but agreement does
not by itself prove the graph is correct.
The statement $N \nindep Y \mid E$ is especially important as a warning
to practitioners: a researcher who conditions only on education when
studying the neighborhood--earnings association will \emph{introduce}
bias through the activated collider path, not remove it.

\medskip
\noindent\textbf{Step 4 --- Identification.}
We wish to identify $P(y \mid \doop(E{=}e))$, the distribution of earnings
under an intervention that sets education to $e$.
The back-door criterion (Chapter~3) requires an adjustment set
$\mathbf{S}$ that (i) blocks every back-door path from $E$ to $Y$ and
(ii) contains no descendant of $E$.

The only back-door path is
\[
  E \leftarrow B \to Y,
\]
a fork at $B$.
Consider two candidate adjustment sets:
\begin{itemize}
  \item $\mathbf{S} = \{B\}$: this blocks the path $E \leftarrow B \to Y$,
        and $B$ is not a descendant of $E$.
        Hence $\{B\}$ is valid.
  \item $\mathbf{S} = \{N\}$: this does not block the path
        $E \leftarrow B \to Y$, so $\{N\}$ is invalid.
\end{itemize}
This illustrates why every back-door path must be blocked: although $N$
is upstream of $E$, conditioning on $N$ does nothing to close the
confounding fork $E \leftarrow B \to Y$.

With $\mathbf{S} = \{B\}$, the back-door formula gives
\[
  P(y \mid \doop(E{=}e))
  \;=\; \sum_{b} P(y \mid e, b)\,P(b).
\]
Every term on the right-hand side involves only the observed distribution,
so the causal effect is identified entirely from observational data ---
the payoff of the graphical analysis.

% ==============================================================
\section{The Big Picture}
% ==============================================================

The concepts developed in this chapter form the graphical foundation for
the identification and estimation theory developed later in the course.
The central logic runs from a qualitative causal graph to an identification
formula for an interventional quantity:

\bigskip
\begin{center}
\begin{tikzpicture}[
    box/.style = {
      rectangle, rounded corners=6pt,
      draw=isublue, fill=defbg, thick,
      text width=7.2cm, align=center,
      minimum height=1.1cm,
      font=\bfseries\color{isublue}
    },
    tool/.style = {
      rectangle, rounded corners=3pt,
      draw=accent, fill=white, thin,
      text width=4.6cm, align=center,
      font=\small\itshape\color{accent}
    },
    arr/.style  = {-{Stealth[length=7pt,width=5pt]}, line width=1.6pt, color=accent},
    node distance = 0.55cm
  ]
  % Four main boxes
  \node[box] (G)  {Graph Structure};
  \node[box, below=1.2cm of G]  (D)  {d-Separation};
  \node[box, below=1.2cm of D]  (CI) {Conditional Independence (CI) };
  \node[box, below=1.2cm of CI] (ID) {Identification Formula};

  % Arrows
  \draw[arr] (G)  -- (D);
  \draw[arr] (D)  -- (CI);
  \draw[arr] (CI) -- (ID);

  % Side annotations
  \node[tool, right=1.1cm of G]
    {DAG encodes direct causal relationships};
  \node[tool, right=1.1cm of D]
    {Read off all $\indep$ statements\\ from the graph};
  \node[tool, right=1.1cm of CI]
    {Interpret or assess implied\\ independences in data};
  \node[tool, right=1.1cm of ID]
    {Back-door / front-door /\\ do-calculus (Chapter~3)};

  % Horizontal connectors (dotted)
  \foreach \n in {G, D, CI, ID}
    \draw[dotted, color=accent!60, thin] (\n.east) -- ++( 1.1cm, 0);
\end{tikzpicture}
\end{center}
\bigskip

The figure summarizes the workflow visually; the paragraph below states
the same logic in words.
Each arrow represents a distinct step in causal reasoning.
First, we encode substantive assumptions in a DAG\@.
Second, we use d-separation to read off the conditional independence
structure implied by that graph.
Third, under the Markov property, we translate those graphical statements
into probabilistic restrictions.
Fourth, we use those restrictions to derive identification formulas for
interventional quantities such as $P(y \mid \doop(t))$.
Section~\ref{w2:sec:education} illustrates the first three steps in full
and previews the fourth through the back-door formula.

In more abstract form, the logic of the chapter is
\[
  \text{structural assumptions}
  \;\Longrightarrow\;
  \text{DAG}
  \;\Longrightarrow\;
  \text{d-separation}
  \;\Longrightarrow\;
  \text{CI under the Markov property}
  \;\Longrightarrow\;
  \text{identification formulas in later chapters}.
\]
Chapter~2 establishes the first four links in this chain.
Chapters~3 and beyond use the same graphical machinery to derive specific
identification results, including back-door adjustment, front-door
adjustment, and do-calculus formulas.

% ==============================================================
\section{Summary}
% ==============================================================

\begin{enumerate}
  \item \textbf{DAGs as causal structure.}
        A DAG is a directed acyclic graph whose nodes represent variables
        and whose directed edges represent direct causal relationships.
        Once a DAG is specified, it encodes which variables are causally
        connected, which variables are potential confounders, and which
        variables may block or open paths.

  \item \textbf{Three-node motifs.}
        Every path is built from three local structures: chains, forks,
        and colliders.  Chains and forks are open by default and are
        blocked by conditioning on the middle node.  Colliders are blocked
        by default and are opened by conditioning on the collider or on
        one of its descendants.

  \item \textbf{d-Separation and conditional independence.}
        The d-separation criterion is a graphical rule: $X$ and $Y$ are
        d-separated by a set $\mathbf{S}$ exactly when every path between
        them is blocked by $\mathbf{S}$.  Under the Markov property, a
        d-separation statement implies the corresponding conditional
        independence in the distribution.  Some such implications may be
        assessed empirically, but agreement with the data does not by
        itself verify the graph, and assumptions involving unobserved
        variables are generally not testable.

  \item \textbf{Markov factorization.}
        The local Markov property yields the factorization
        \[
          p(v_1,\dots,v_k)
          \;=\;
          \prod_{i=1}^{k} p\!\left(v_i \mid \Pa(v_i)\right),
        \]
        which expresses the joint distribution in terms of local
        conditional distributions.  This factorization is the bridge from
        graphical structure to probability calculus and underlies later
        identification arguments.

  \item \textbf{Collider bias.}
        Conditioning on a collider --- or on a descendant of a collider
        --- can induce a spurious association between otherwise independent
        variables \citep{berkson1946limitations}.  This is the opposite of
        confounding adjustment: conditioning on a common cause removes
        spurious association, whereas conditioning on a common effect
        creates it.

  \item \textbf{A practical workflow.}
        To analyze a DAG, proceed in four steps: identify the graph
        structure, determine which paths are open or blocked, translate
        d-separation statements into conditional independences under the
        Markov property, and then interpret those independences in light
        of the causal question.  This workflow will be reused throughout
        the rest of the course.
\end{enumerate}

% ==============================================================
\newpage
\section*{Problems}
\addcontentsline{toc}{section}{Problems}
% ==============================================================

\begin{enumerate}[label=\textbf{\arabic*.}, leftmargin=*, itemsep=10pt]

  \item \textbf{Warm-up: a single collider.}
  Consider the DAG $X \to Y \leftarrow Z$, where $X$ and $Z$ have no
  other connections.
  \begin{enumerate}[label=(\alph*)]
    \item Identify the structural role of $Y$ on the path $X \to Y \leftarrow Z$.
    \item Is $X \indep Z$?  Apply the d-separation criterion to
          the only path between $X$ and $Z$, and state which blocking rule
          applies.
    \item Is $X \indep Z \mid Y$?  Explain what happens to the
          path when $Y$ is conditioned on, and describe in one sentence the
          real-world phenomenon this illustrates.
  \end{enumerate}

  \item \textbf{d-Separation practice.}
  Consider the DAG: $A \to B \to D$, $A \to C \to D$, $B \to E$, $C \to E$.
  \begin{enumerate}[label=(\alph*)]
    \item List all paths between $A$ and $E$.  (\emph{Hint}: there are four
          paths in total; two pass through $D$.)
    \item For each path, identify the role (chain, fork, collider) of each
          intermediate node.
    \item Does $\{B, C\}$ d-separate $A$ and $E$?
    \item Does $\{D\}$ d-separate $B$ and $C$? What type of node is $D$
          on the path $B \to D \leftarrow C$?
  \end{enumerate}

  \item \textbf{Markov factorization and collider activation.}
  Consider the DAG with edges $A \to E$, $A \to W$, $F \to E$, $E \to W$,
  where $A$ = ability, $F$ = family income, $E$ = education, $W$ = wages.
  \begin{enumerate}[label=(\alph*)]
    \item Write down the Markov factorization $p(a, f, e, w)$.
    \item Is $(F \indep W)_{\Gcal}$?
          List all paths between $F$ and $W$ and determine which are open.
    \item Is $(F \indep W \mid E)_{\Gcal}$?
          Identify the role of $E$ on each path and explain whether
          conditioning on $E$ opens or closes each one.
    \item A researcher regresses $W$ on $E$ and $F$, omitting $A$.
          Is the coefficient on $E$ a causal effect of education on wages?
          Explain using the graph.
  \end{enumerate}

  \item \textbf{Theorem~\ref{w2:thm:dsep} for the collider.}
  Consider the collider $A \to M \leftarrow B$ with Markov factorization
  $p(a, m, b) = p(a)\,p(b)\,p(m \mid a, b)$.
  \begin{enumerate}[label=(\alph*)]
    \item By marginalising over $M$, show that $A \indep B$
          in the joint distribution (i.e., $p(a,b) = p(a)\,p(b)$).
          This verifies Theorem~\ref{w2:thm:dsep} for the case
          $\mathbf{S} = \varnothing$, where
          Example~\ref{w2:ex:threesettings}(iii) established d-separation
          graphically.
    \item Show that conditioning on $M$ breaks this independence: write
          out $p(a, b \mid m)$ and explain why it does \emph{not} factorise
          into $p(a \mid m)\,p(b \mid m)$ in general.  What does this imply
          about $A$ and $B$ among hospitalised patients in the
          accident--hospitalization--cancer example?
    \item Explain in one sentence why parts~(a) and~(b) together are
          consistent with Theorem~\ref{w2:thm:dsep}.  (\emph{Hint}:
          Theorem~\ref{w2:thm:dsep} is a one-directional statement.)
  \end{enumerate}

  \item \textbf{Terminology check.}
  Consider the DAG with edges $U \to X$, $U \to Z$, $X \to W$,
  $Z \to W$, $W \to Y$.
  \begin{enumerate}[label=(\alph*)]
    \item Identify the parents, children, ancestors, descendants, and
          non-descendants of node $W$.
    \item List all pairs of adjacent nodes (connected by a single edge).
    \item Which pairs of nodes are connected by a directed path?
          List every such pair and the corresponding directed path.
    \item Write the Markov factorization $p(u, x, z, w, y)$ implied by
          this DAG.
  \end{enumerate}

  \item \textbf{Markov factorization and local Markov property.}
  Consider the DAG with edges $X_1 \to X_2$, $X_1 \to X_3$,
  $X_2 \to X_4$, $X_3 \to X_4$.
  \begin{enumerate}[label=(\alph*)]
    \item Write the joint density $p(x_1, x_2, x_3, x_4)$ implied by the
          Markov factorization.
    \item State the local Markov property for each of the four nodes.
          For each node, identify the conditioning set $\Pa(X_i)$ and
          the independence set $\Nd(X_i) \setminus \Pa(X_i)$.
    \item Is $(X_2 \indep X_3)_{\Gcal}$?  Is $(X_2 \indep X_3 \mid X_1)_{\Gcal}$?
          Justify each answer by listing all paths between $X_2$ and $X_3$
          and checking whether each is blocked.
  \end{enumerate}

  \item \textbf{Toy proof: conditional independence in the fork.}
  Consider the fork $X_1 \leftarrow X_2 \to X_3$ with Markov factorization
  $p(x_1, x_2, x_3) = p(x_2)\,p(x_1 \mid x_2)\,p(x_3 \mid x_2)$.
  \begin{enumerate}[label=(\alph*)]
    \item Show directly, by conditioning on $X_2 = x_2$, that
          $X_1 \indep X_3 \mid X_2$.  Follow the same three steps used in
          Example~\ref{w2:ex:chainproof} for the chain.
    \item Is $X_1 \indep X_3$ marginally?  Justify your answer both
          graphically (using d-separation) and algebraically (by
          marginalising over $X_2$).
    \item Explain in one sentence what the fork represents substantively
          and why conditioning on the common cause $X_2$ removes the
          association between $X_1$ and $X_3$.
  \end{enumerate}

  \item \textbf{d-Separation in the practice DAG.}
  Refer to the eight-node DAG in Example~\ref{w2:ex:practicedag}
  (nodes $W$, $X_1$, $T$, $M_1$, $M_2$, $Y$, $C$, $D$;
  edges $W \to X_1$, $W \to T$, $W \to Y$, $T \to M_1$, $M_1 \to Y$,
  $M_1 \to M_2$, $X_1 \to C$, $M_2 \to C$, $C \to D$).
  For each query below, state whether it is true or false and justify
  your answer by listing all relevant paths and determining whether each
  is blocked or open.  Do not refer to the worked solutions in
  Example~\ref{w2:ex:practicedag} until you have written your own answer.
  \begin{enumerate}[label=(\alph*)]
    \item $X_1 \indep Y$
    \item $X_1 \indep Y \mid W$
    \item $T \indep M_2 \mid M_1$
    \item Does conditioning on $C$ induce collider bias between $X_1$
          and $M_2$?  Identify the specific path that is opened and explain
          why the induced association is spurious.
    \item Does conditioning on $D$ (a descendant of $C$) open the path
          $X_1 \to C \leftarrow M_2$?  State which clause of
          Definition~\ref{w2:def:dsep} applies.
  \end{enumerate}

\end{enumerate}

% ==============================================================
\newpage
\ifSubfilesClassLoaded{
  \bibliographystyle{plainnat}
  \bibliography{causal_inference}
}{}

\end{document}